---
title: "Data Basics"
editor:
  mode: source
---

## Reading the climate record of Yale-Myers Forest

Yale-Myers Forest is a 7,840 acre research forest in northeastern Connecticut. It is a hub for education, research, and harvesting operations within the Yale Forest School system. The forest primarily consists of mixed hardwoods on glacial till soils, with hemlock, scattered white pine stands from old field origins, and red pine plantations established in the 1940s.

In this workshop, we will use daily climate data to ask a more interesting question:

> Has Yale-Myers Forest become warmer, wetter, or more climatically stressful since 1990?

By the end of the workshop, you will have built a small climate evidence brief for Yale-Myers Forest. Along the way, you will learn to:

1.  Download climate data with `daymetr`.
2.  Manipulate data with `tidyverse`:
    -   Format date elements.
    -   Use `select()` to choose variables from a data frame.
    -   Use `filter()` to choose data based on values.
    -   Use `mutate()` to create new variables.
    -   Use `reframe()` to summarize datasets.
    -   Use `full_join()` to merge datasets.
3.  Visualize data with `ggplot2`.
4.  Write basic functions that make your analysis easier to reuse.

# Data manipulation with tidyverse

### Load the required libraries

```{r}
#| output: false
library(daymetr)
library(tidyverse)
```

### Download daily climate data

To explore Yale-Myers Forest, download daily Daymet data for 1990-2018 at latitude 41.952909 and longitude -72.123859.

Review the arguments for the `download_daymet()` function:

```{r}
#| eval: false
?download_daymet()
```

Download the data:

```{r}
yale_myers_daymet <- download_daymet(lat = 41.952909,
                                      lon = -72.123859,
                                      start = 1990,
                                      end = 2018)
```

You now have a nested list that includes site information and daily climate data. The daily climate data are stored in the `data` element of the list. Pull the data out and convert it to a data frame:

```{r}
yale_myers <- as.data.frame(yale_myers_daymet$data)
```

Look at the data:

```{r}
head(yale_myers)
summary(yale_myers)
```

### Turn daily records into a usable climate timeline

Daymet stores time as `year` and day of year (`yday`). We need one actual date column before we can summarize the data by month, season, or year.

Step 1. Use the year and day of year to create a date:

```{r}
yale_myers$year_day <- paste(yale_myers$year, yale_myers$yday, sep = "-") %>%
  as.Date(format = "%Y-%j")
```

Look at your work:

```{r}
head(yale_myers$year_day)
```

Step 2. Format the date as YYYY-mm-dd:

```{r}
yale_myers$date <- format(yale_myers$year_day, "%Y-%m-%d") %>%
  as.Date()
```

Check the result:

```{r}
head(yale_myers$date)
class(yale_myers$date)
```

### Create climate variables with mutate()

The function `mutate()` creates new columns from existing columns. We will create several columns that help us interpret the climate record:

-   `month`: the calendar month.
-   `month_name`: the month name, used for plotting.
-   `season`: a rough ecological season.
-   `julian_day`: the day of year.
-   `tmean`: mean daily temperature.
-   `hot_day`: days with maximum temperature at or above 30 degrees C.
-   `freeze_day`: days with minimum temperature at or below 0 degrees C.
-   `growing_season`: May through September.

```{r}
yale_myers <- yale_myers %>%
  mutate(month = format(date, format = "%m"),
         month_name = factor(month.abb[as.numeric(month)], levels = month.abb),
         julian_day = format(date, format = "%j"),
         season = case_when(
           month %in% c("12", "01", "02") ~ "Winter",
           month %in% c("03", "04", "05") ~ "Spring",
           month %in% c("06", "07", "08") ~ "Summer",
           month %in% c("09", "10", "11") ~ "Fall"
         ),
         tmean = (0.606 * as.numeric(tmax..deg.c.)) +
           (0.394 * as.numeric(tmin..deg.c.)),
         hot_day = tmax..deg.c. >= 30,
         freeze_day = tmin..deg.c. <= 0,
         growing_season = month %in% c("05", "06", "07", "08", "09"))
```

Look at the new variables:

```{r}
yale_myers %>%
  select(date, month, month_name, season, tmean, hot_day, freeze_day, growing_season) %>%
  head()
```

### Rename columns

Use `rename()` to give the Daymet columns shorter names:

```{r}
yale_myers_rn <- yale_myers %>%
  rename(tmax = tmax..deg.c.,
         tmin = tmin..deg.c.,
         prcp = prcp..mm.day.)
```

Look at your work:

```{r}
head(yale_myers_rn)
```

### Select the variables needed for the investigation

Subset the dataset to keep only the variables we need:

```{r}
yale_myers_sub <- yale_myers_rn %>%
  select(date, month, month_name, season, year, julian_day,
         tmax, tmin, tmean, prcp, hot_day, freeze_day, growing_season)
```

Look at your work:

```{r}
head(yale_myers_sub)
```

### Summarize climate conditions with reframe()

Create an annual climate summary. Each row is one year of evidence:

```{r}
yale_myers_annual <- yale_myers_sub %>%
  reframe(.by = year,
          tmax = max(tmax, na.rm = TRUE),
          tmin = min(tmin, na.rm = TRUE),
          tmean = mean(tmean, na.rm = TRUE),
          prcp = sum(prcp, na.rm = TRUE),
          hot_days = sum(hot_day, na.rm = TRUE),
          freeze_days = sum(freeze_day, na.rm = TRUE),
          wettest_day = max(prcp, na.rm = TRUE))
```

Look at your work:

```{r}
head(yale_myers_annual)
```

Determine average monthly conditions. This gives us the seasonal climate fingerprint of Yale-Myers Forest:

```{r}
yale_myers_monthly <- yale_myers_sub %>%
  reframe(.by = c(month, month_name),
          tmax = max(tmax, na.rm = TRUE),
          tmin = min(tmin, na.rm = TRUE),
          tmean = mean(tmean, na.rm = TRUE),
          prcp = sum(prcp, na.rm = TRUE) / length(unique(year)),
          hot_days = sum(hot_day, na.rm = TRUE) / length(unique(year)),
          freeze_days = sum(freeze_day, na.rm = TRUE) / length(unique(year))) %>%
  arrange(month)
```

Look at your work:

```{r}
head(yale_myers_monthly)
```

Determine average conditions by both year and month:

```{r}
yale_myers_yearmon <- yale_myers_sub %>%
  reframe(.by = c(year, month, month_name),
          tmax = max(tmax, na.rm = TRUE),
          tmin = min(tmin, na.rm = TRUE),
          tmean = mean(tmean, na.rm = TRUE),
          prcp = sum(prcp, na.rm = TRUE),
          hot_days = sum(hot_day, na.rm = TRUE),
          freeze_days = sum(freeze_day, na.rm = TRUE)) %>%
  arrange(year, month)
```

Look at your work:

```{r}
head(yale_myers_yearmon)
```

Summarize growing season conditions by year:

```{r}
yale_myers_growing <- yale_myers_sub %>%
  filter(growing_season) %>%
  reframe(.by = year,
          growing_tmean = mean(tmean, na.rm = TRUE),
          growing_prcp = sum(prcp, na.rm = TRUE),
          growing_hot_days = sum(hot_day, na.rm = TRUE))
```

Look at your work:

```{r}
head(yale_myers_growing)
```

### Select a file for later comparison

Subset mean annual temperature and year from the annual summary. Then rename `tmean` to `yale_myers`:

```{r}
yale_myers_annual_tmean <- yale_myers_annual %>%
  mutate(yale_myers = tmean) %>%
  select(year, yale_myers)
```

Look at your work:

```{r}
head(yale_myers_annual_tmean)
```

### Save your work

Save the three files `yale_myers`, `yale_myers_sub`, and `yale_myers_annual_tmean` in a `.RDATA` object. You will need this for the post-workshop assessment:

```{r}
#| eval: false
save(yale_myers, yale_myers_sub, yale_myers_annual_tmean,
     file = "data/products/data_basics.RDATA")
```

# Introduction to ggplot2

`ggplot2` is a plotting package in R. Graphics are built layer by layer. In this workshop, each plot should help answer part of the climate question instead of simply demonstrating a plotting command.

Load the library:

```{r}
library(ggplot2)
```

### Building blocks of the grammar of graphics

-   Data: the dataset.
-   Aesthetics: how variables map to visual features such as x-axis, y-axis, color, fill, size, labels, alpha, shape, line width, and line type.
-   Geometrics: how the data are displayed, such as points, lines, histograms, bars, or boxplots.
-   Facets: split the data into subsets.
-   Statistics: binning, smoothing, summaries, or model fits.
-   Coordinates: the space between data and display.
-   Themes: non-data visual elements.

### Data layer

We first define the source of the information to be visualized:

```{r}
ggplot(data = yale_myers_annual) +
  labs(title = "Annual climate record of Yale-Myers Forest")
```

### Aesthetic layer

The aesthetic layer maps variables to visual features. Here, `year` is mapped to the x-axis and annual mean temperature is mapped to the y-axis:

```{r}
ggplot(data = yale_myers_annual,
       aes(x = year, y = tmean)) +
  labs(title = "Annual mean temperature at Yale-Myers Forest")
```

### Geometric layer

The geometric layer controls how the data are drawn:

```{r}
ggplot(data = yale_myers_annual,
       aes(x = year, y = tmean)) +
  geom_point() +
  geom_line() +
  labs(title = "Annual mean temperature at Yale-Myers Forest",
       x = "Year",
       y = "Mean air temperature (degrees C)")
```

Add size and color to show whether warmer years also had more hot days:

```{r}
ggplot(data = yale_myers_annual,
       aes(x = year, y = tmean, size = hot_days, color = hot_days)) +
  geom_point() +
  labs(title = "Warm years and hot days at Yale-Myers Forest",
       x = "Year",
       y = "Mean air temperature (degrees C)",
       size = "Hot days",
       color = "Hot days")
```

Create a histogram to see the distribution of daily mean temperature:

```{r}
ggplot(data = yale_myers_sub,
       aes(x = tmean)) +
  geom_histogram(binwidth = 5) +
  labs(title = "Distribution of daily mean temperature",
       x = "Daily mean air temperature (degrees C)",
       y = "Count")
```

### Seasonal climate fingerprint

Use the monthly summary to see the shape of a typical year:

```{r}
ggplot(data = yale_myers_monthly,
       aes(x = month_name, y = tmean, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Seasonal temperature pattern at Yale-Myers Forest",
       x = "Month",
       y = "Mean air temperature (degrees C)")
```

Now look at precipitation:

```{r}
ggplot(data = yale_myers_monthly,
       aes(x = month_name, y = prcp)) +
  geom_col() +
  labs(title = "Average monthly precipitation at Yale-Myers Forest",
       x = "Month",
       y = "Average monthly precipitation (mm)")
```

### Facet layer

The facet layer splits the data into subsets. Here we compare fall months, which are often important for field work and forest management:

```{r}
yale_myers_fall <- yale_myers_sub %>%
  filter(month > "08" & month < "12")

p <- ggplot(data = yale_myers_fall,
            aes(x = date, y = tmean)) +
  geom_point()

p + facet_grid(month_name ~ .) +
  labs(title = "Fall temperature records at Yale-Myers Forest",
       x = "Date",
       y = "Mean air temperature (degrees C)")
```

### Statistics layer

We can add a model fit to ask whether a seasonal climate variable is changing over time. Here we explore February temperatures from 1990-2018:

```{r}
yale_myers_feb <- yale_myers_sub %>%
  filter(month == "02")

ggplot(data = yale_myers_feb,
       aes(x = date, y = tmean)) +
  geom_point() +
  stat_smooth(method = lm, col = "red") +
  labs(title = "Are February temperatures changing at Yale-Myers Forest?",
       x = "Date",
       y = "Mean air temperature (degrees C)")
```

Use the annual summary to ask whether hot days are changing:

```{r}
ggplot(data = yale_myers_annual,
       aes(x = year, y = hot_days)) +
  geom_col() +
  stat_smooth(method = lm, se = FALSE, col = "red") +
  labs(title = "Annual number of hot days at Yale-Myers Forest",
       x = "Year",
       y = "Number of days with tmax >= 30 degrees C")
```

### Coordinates layer

Coordinates control the plotted space. First, plot fall minimum temperature against mean temperature:

```{r}
ggplot(data = yale_myers_fall,
       aes(x = tmin, y = tmean)) +
  geom_point() +
  stat_smooth(method = lm, col = "red") +
  scale_y_continuous(limits = c(-10, 40), expand = c(0, 0)) +
  scale_x_continuous(limits = c(-10, 40), expand = c(0, 0)) +
  coord_equal() +
  labs(title = "Fall minimum and mean temperature",
       x = "Minimum air temperature (degrees C)",
       y = "Mean air temperature (degrees C)")
```

Use `coord_cartesian()` to zoom without deleting data outside the visible range:

```{r}
ggplot(data = yale_myers_fall,
       aes(x = tmin, y = tmean)) +
  geom_point() +
  stat_smooth(method = lm, col = "red") +
  labs(title = "Zoomed view of fall temperature relationships",
       x = "Minimum air temperature (degrees C)",
       y = "Mean air temperature (degrees C)") +
  coord_cartesian(xlim = c(-10, 10))
```

### Theme layer

The theme layer controls the finer points of display, such as font size, background color, and grid lines:

```{r}
ggplot(data = yale_myers_fall,
       aes(x = tmin, y = tmean)) +
  geom_point() +
  stat_smooth(method = lm, col = "red") +
  facet_grid(. ~ month_name) +
  theme(plot.background = element_rect(fill = "lightblue", colour = "gray")) +
  labs(title = "Fall temperature relationships by month",
       x = "Minimum air temperature (degrees C)",
       y = "Mean air temperature (degrees C)")
```

# Comparing Yale-Myers Forest to New Haven

Yale-Myers Forest is inland and forested. New Haven is coastal and urban. Comparing the two locations gives `full_join()` a purpose: we can ask whether the same years were warm at both locations.

Download Daymet data for New Haven:

```{r}
new_haven_daymet <- download_daymet(lat = 41.3083,
                                     lon = -72.9279,
                                     start = 1990,
                                     end = 2018)

new_haven <- as.data.frame(new_haven_daymet$data)
```

Prepare the New Haven data using the same steps:

```{r}
new_haven_sub <- new_haven %>%
  mutate(date = paste(year, yday, sep = "-") %>%
           as.Date(format = "%Y-%j"),
         month = format(date, format = "%m"),
         tmean = (0.606 * as.numeric(tmax..deg.c.)) +
           (0.394 * as.numeric(tmin..deg.c.))) %>%
  rename(tmax = tmax..deg.c.,
         tmin = tmin..deg.c.,
         prcp = prcp..mm.day.) %>%
  select(date, month, year, tmax, tmin, tmean, prcp)
```

Create a mean annual temperature summary for New Haven:

```{r}
new_haven_annual_tmean <- new_haven_sub %>%
  reframe(.by = year,
          new_haven = mean(tmean, na.rm = TRUE))
```

Use `full_join()` to combine the annual temperature summaries:

```{r}
annual_tmean_compare <- yale_myers_annual_tmean %>%
  full_join(new_haven_annual_tmean, by = "year")
```

Look at the joined data:

```{r}
head(annual_tmean_compare)
```

Visualize the comparison:

```{r}
annual_tmean_compare_long <- annual_tmean_compare %>%
  pivot_longer(cols = c(yale_myers, new_haven),
               names_to = "location",
               values_to = "tmean")

ggplot(data = annual_tmean_compare_long,
       aes(x = year, y = tmean, color = location)) +
  geom_line() +
  geom_point() +
  labs(title = "Annual mean temperature comparison",
       x = "Year",
       y = "Mean air temperature (degrees C)",
       color = "Location")
```

# Introduction to writing functions in R

Functions are reusable sets of instructions. They help you avoid repeating code and make your analysis easier to check. In environmental data science, functions are especially useful when you need to apply the same summary to multiple locations, variables, or time periods.

### What is a function?

In R, a function is an object that performs a task. R comes with built-in functions, such as:

-   `min()`, `max()`, `mean()`, and `median()`.
-   `sum()`.
-   `range()`.
-   `str()`.
-   `print()`.
-   `length()`.
-   `sort()`.

You can also create your own functions. The basic syntax is:

```{r}
#| eval: false
function_name <- function(parameters) {
  # code to run goes here
}
```

### Write a standard error function

The standard error measures the precision of an estimated mean. We will use this function in the assignment:

```{r}
standard_error <- function(x) {
  sd(x, na.rm = TRUE) / sqrt(sum(!is.na(x)))
}
```

Test the function:

```{r}
standard_error(yale_myers_sub$tmean)
```

### Write a reusable monthly climate summary

Now write a function that summarizes monthly climate conditions for any Daymet-style data frame that already has `month`, `tmean`, `prcp`, `tmin`, and `tmax` columns:

```{r}
summarize_monthly_climate <- function(data) {
  data %>%
    reframe(.by = month,
            tmean = mean(tmean, na.rm = TRUE),
            tmean_se = standard_error(tmean),
            prcp = mean(prcp, na.rm = TRUE),
            prcp_se = standard_error(prcp),
            tmin = mean(tmin, na.rm = TRUE),
            tmin_se = standard_error(tmin),
            tmax = mean(tmax, na.rm = TRUE),
            tmax_se = standard_error(tmax)) %>%
    arrange(month)
}
```

Use the function on both locations:

```{r}
yale_myers_monthly_summary <- summarize_monthly_climate(yale_myers_sub)
new_haven_monthly_summary <- summarize_monthly_climate(new_haven_sub)
```

Add location labels and combine the data:

```{r}
monthly_location_compare <- bind_rows(
  yale_myers_monthly_summary %>% mutate(location = "Yale-Myers Forest"),
  new_haven_monthly_summary %>% mutate(location = "New Haven")
)
```

Plot monthly mean temperature with standard error:

```{r}
ggplot(data = monthly_location_compare,
       aes(x = month, y = tmean, color = location, group = location)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = tmean - tmean_se,
                    ymax = tmean + tmean_se),
                width = 0.2) +
  labs(title = "Monthly mean temperature with standard error",
       x = "Month",
       y = "Mean air temperature (degrees C)",
       color = "Location")
```

### Function arguments by position and by name

Function arguments can be passed by position:

```{r}
subtract_two_nums <- function(x, y) {
  x - y
}

print(subtract_two_nums(3, 1))
```

Or by name:

```{r}
print(subtract_two_nums(x = 3, y = 1))
print(subtract_two_nums(y = 1, x = 3))
```

When functions become more complex, named arguments often make your code easier to read.

::: {.callout-tip}
## Course product
This workshop produces a climate evidence brief. Use the standard deliverable shape from the Final Project Roadmap: one clear question, one reproducibility note, at least one table or summary, one or two figures, one interpretation paragraph, and one limitation or next-step paragraph.
:::

# Assignment

Create a short climate evidence brief comparing Yale-Myers Forest with one U.S. location of your choice.

Find a location of interest within the United States in Google Earth (<https://earth.google.com/>) and record the latitude and longitude in decimal degrees. Download Daymet data for that location. Use the same workflow from this workshop to clean the data, summarize climate by month, and compare it to Yale-Myers Forest.

Your brief should include figures and text that show:

1.  Differences in monthly mean temperature (`tmean`) and standard error for both locations.
2.  Differences in monthly precipitation (`prcp`) and standard error for both locations.
3.  Differences in monthly minimum temperature (`tmin`) and standard error for both locations.
4.  Differences in monthly maximum temperature (`tmax`) and standard error for both locations.
5.  One climate stress indicator of your choice, such as hot days, freeze days, or wettest day.
6.  A short explanation of why the two locations are similar or different.

Create a report to upload to Canvas and save it as a PDF. Consider creating this report in R Markdown or Quarto: <https://hbctraining.github.io/Training-modules/Rmarkdown/lesson.html>
