---
title: "Data Integration: Building a Methane Flux Modeling Table"
editor:
  mode: source
---

## From data sources to a model-ready table

The final course project asks you to build a methane flux model. Before you can fit a model, you need to compile a dataset where each row is a clean observation and each column is a response variable, predictor, or useful descriptor.

In this workshop, we will ask:

> How do we combine flux measurements, site metadata, vegetation indices, soil, climate, elevation, and reanalysis data into one table that can be used to model methane flux?

The target data product is a **site-month modeling table**:

-   one row = one tower site in one month;
-   response = methane flux (`FCH4_F_gC`);
-   dynamic predictors = monthly tower climate, MODIS vegetation indices, ERA5 climate;
-   static predictors = ecosystem type, country, soil, climate normals, elevation.

By the end of this workshop, you will be able to:

1.  Identify whether a data source is tabular, vector, or raster.
2.  Identify whether a predictor is static or dynamic.
3.  Choose join keys for site-only and site-time datasets.
4.  Extract raster values to tower locations.
5.  Join static site attributes to monthly observations.
6.  Join dynamic monthly predictors by `SITE_ID` and `YearMon`.
7.  Check joins for missing values, duplicate keys, and row-count changes.
8.  Save a model-ready methane flux dataset.

# Data inventory

Before writing code, make a data inventory. Every predictor you use in a model should have a source, units, spatial/temporal resolution, and join key.

| Source | Data Type | Static or Dynamic? | What It Adds | Join Key |
|---|---|---|---|---|
| `FluxNet_Sites_2024.csv` | Tabular points | Static | site location, ecosystem type, site MAT/MAP | `SITE_ID` |
| `GlobalSoil_grids.tif` | Raster | Static | soil bulk density, pH, nitrogen | location |
| `GlobalClimate.tif` | Raster | Static | climate normals | location |
| `Elevation.tif` | Raster | Static | elevation | location |
| `ENV720-MOD13A3-061-results.csv` | Tabular remote sensing | Dynamic monthly | NDVI, EVI | `SITE_ID`, `YearMon` |
| `FLUXNET_FLUXES.RDATA` | Tabular flux | Dynamic monthly | methane flux response, tower meteorology | `SITE_ID`, `YearMon` |
| ERA5 rasters | Raster reanalysis | Dynamic monthly | gridded precipitation and air temperature | location, `YearMon` |

::: {.callout-note}
## Static versus dynamic predictors
Static predictors describe a site and can be repeated across months. Dynamic predictors change through time and must be joined by both site and time. Accidentally joining monthly data only by `SITE_ID` is one of the easiest ways to create duplicated or incorrect model data.
:::

# Getting a Climate Data Store (CDS) API key

Later in this workshop, we will use ERA5 monthly climate data from the Copernicus Climate Data Store (CDS). This requires a free personal account and API key.

1.  Create a free account at <https://cds.climate.copernicus.eu/>.
2.  Once logged in, open your user profile page and copy your **Personal Access Token**.
3.  Register it once per computer:

```{r}
#| eval: false
install.packages("ecmwfr")
library(ecmwfr)
wf_set_key(key = "paste-your-personal-access-token-here")
```

::: {.callout-warning}
## Never save your API key in a script
Type your key directly into the R console, not into a `.qmd` file you save or push to GitHub. Treat it like a password.
:::

# Load libraries

```{r}
#| output: false
library(sf)
library(terra)
library(tidyverse)
library(ggplot2)
library(rnaturalearth)
library(zoo)
library(lubridate)
```

# Helper checks for joins

We will use a few small checks repeatedly. These are habits you should use in your final project.

```{r}
count_missing <- function(data) {
  tibble(variable = names(data),
         missing_count = vapply(data, function(x) sum(is.na(x)), integer(1))) %>%
    arrange(desc(missing_count))
}
```

Check for duplicate keys:

```{r}
find_duplicate_keys <- function(data, ...) {
  data %>%
    count(..., name = "n") %>%
    filter(n > 1)
}
```

# Static site table

Import the FLUXNET site table:

```{r}
fluxnet_sites <- read.csv("data/FluxNet_Sites_2024.csv")
```

Keep sites with methane data and valid coordinates:

```{r}
fluxnet_ch4 <- fluxnet_sites %>%
  filter(FLUXNET.CH4 != "",
         !is.na(LOCATION_LONG),
         !is.na(LOCATION_LAT))

nrow(fluxnet_sites)
nrow(fluxnet_ch4)
```

Convert the methane site table to a simple feature object:

```{r}
fluxnet_ch4_sf <- st_as_sf(x = fluxnet_ch4,
                            coords = c("LOCATION_LONG", "LOCATION_LAT"),
                            crs = 4326,
                            remove = FALSE)

ggplot(data = fluxnet_ch4_sf) +
  geom_sf() +
  labs(title = "FLUXNET CH4 tower locations")
```

# Add country and continent

Country and continent are static site attributes. We get them by spatially joining tower points to country polygons.

```{r}
world_sf <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent %in% c("Europe", "Asia", "North America",
                          "South America", "Africa", "Oceania")) %>%
  st_make_valid()

world_context <- world_sf %>%
  select(country = name, continent, geometry)

fluxnet_ch4_sf <- fluxnet_ch4_sf %>%
  st_join(world_context, join = st_within)
```

Check the new attributes:

```{r}
fluxnet_ch4_sf %>%
  st_drop_geometry() %>%
  select(SITE_ID, SITE_NAME, country, continent) %>%
  head()
```

# Extract static raster predictors

Static raster predictors are joined by location. We will extract soil, climate normals, and elevation to each tower point.

### Soil

```{r}
soil <- terra::rast("data/GlobalSoil_grids.tif")
names(soil)
```

Transform the tower points to the raster CRS, then extract:

```{r}
fluxnet_ch4_soil <- st_transform(fluxnet_ch4_sf, crs = crs(soil))

soil_values <- terra::extract(soil, vect(fluxnet_ch4_soil)) %>%
  select(-ID) %>%
  rename(soil_bulk_density = BulkDensity,
         soil_ph = PH,
         soil_nitrogen = Nitrogen)
```

### Climate normals

```{r}
climate <- terra::rast("data/GlobalClimate.tif")
names(climate)
```

```{r}
fluxnet_ch4_climate <- st_transform(fluxnet_ch4_sf, crs = crs(climate))

climate_values <- terra::extract(climate, vect(fluxnet_ch4_climate)) %>%
  select(-ID) %>%
  rename(climate_map = MAP,
         climate_tmin = TMIN,
         climate_tmax = TMAX,
         climate_mat = MAT)
```

### Elevation

```{r}
elevation <- terra::rast("data/Elevation.tif")
names(elevation)
```

```{r}
fluxnet_ch4_elevation <- st_transform(fluxnet_ch4_sf, crs = crs(elevation))

elevation_values <- terra::extract(elevation, vect(fluxnet_ch4_elevation)) %>%
  select(-ID) %>%
  rename(elevation_m = wc2.1_2.5m_elev)
```

### Build the static site table

Combine the site metadata and static raster values:

```{r}
site_static <- fluxnet_ch4_sf %>%
  st_drop_geometry() %>%
  bind_cols(soil_values, climate_values, elevation_values)
```

Keep one row per site:

```{r}
site_static <- site_static %>%
  distinct(SITE_ID, .keep_all = TRUE)

find_duplicate_keys(site_static, SITE_ID)
```

Inspect missing values:

```{r}
site_static %>%
  select(SITE_ID, IGBP, country, continent,
         soil_bulk_density, soil_ph, soil_nitrogen,
         climate_map, climate_mat, elevation_m) %>%
  count_missing() %>%
  head(15)
```

# Dynamic monthly vegetation predictors

MODIS NDVI and EVI are monthly predictors. Their join key must include both `SITE_ID` and `YearMon`.

```{r}
modis_vi <- read.csv("data/ENV720-MOD13A3-061-results.csv")
names(modis_vi)
```

Prepare the MODIS table:

```{r}
fluxnet_vi_monthly <- modis_vi %>%
  select(ID,
         Date,
         MOD13A3_061__1_km_monthly_EVI,
         MOD13A3_061__1_km_monthly_NDVI,
         MOD13A3_061__1_km_monthly_VI_Quality) %>%
  rename(SITE_ID = ID,
         EVI = MOD13A3_061__1_km_monthly_EVI,
         NDVI = MOD13A3_061__1_km_monthly_NDVI,
         QAQC = MOD13A3_061__1_km_monthly_VI_Quality) %>%
  mutate(date_formatted = as.Date(Date),
         YearMon = zoo::as.yearmon(date_formatted)) %>%
  filter(QAQC > 0) %>%
  distinct(SITE_ID, YearMon, .keep_all = TRUE)
```

Check the keys:

```{r}
find_duplicate_keys(fluxnet_vi_monthly, SITE_ID, YearMon)
range(fluxnet_vi_monthly$YearMon)
```

# Monthly flux data

Load the monthly flux dataset. This contains the response variable and tower-measured monthly meteorology.

```{r}
load("data/FLUXNET_FLUXES.RDATA")

fluxnet_flux <- FLUXNET.flux
```

Create `SITE_ID` and inspect the time range:

```{r}
fluxnet_flux <- fluxnet_flux %>%
  mutate(SITE_ID = SITE)

range(fluxnet_flux$YearMon)
names(fluxnet_flux)
```

Check whether `SITE_ID + YearMon` uniquely identifies rows:

```{r}
find_duplicate_keys(fluxnet_flux, SITE_ID, YearMon)
```

# Join dynamic predictors

Start with the flux data, then join MODIS by `SITE_ID` and `YearMon`.

```{r}
before_rows <- nrow(fluxnet_flux)

flux_month_vi <- fluxnet_flux %>%
  left_join(fluxnet_vi_monthly, by = c("SITE_ID", "YearMon"))

after_rows <- nrow(flux_month_vi)

before_rows
after_rows
```

Check for flux rows that did not receive MODIS data:

```{r}
flux_month_vi %>%
  summarize(missing_ndvi = sum(is.na(NDVI)),
            missing_evi = sum(is.na(EVI)),
            total_rows = n())
```

Use `anti_join()` to inspect unmatched keys:

```{r}
fluxnet_flux %>%
  anti_join(fluxnet_vi_monthly, by = c("SITE_ID", "YearMon")) %>%
  count(SITE_ID, sort = TRUE) %>%
  head()
```

# Join static site predictors to monthly rows

Static site attributes join by `SITE_ID` only. This repeats each site's static values across all months for that site.

```{r}
site_static_for_join <- site_static %>%
  select(SITE_ID, SITE_NAME, IGBP, country, continent,
         LOCATION_LAT, LOCATION_LONG, LOCATION_ELEV,
         soil_bulk_density, soil_ph, soil_nitrogen,
         climate_map, climate_tmin, climate_tmax, climate_mat,
         elevation_m)
```

Join site attributes to the monthly table:

```{r}
before_rows <- nrow(flux_month_vi)

flux_model_base <- flux_month_vi %>%
  left_join(site_static_for_join, by = "SITE_ID")

after_rows <- nrow(flux_model_base)

before_rows
after_rows
```

Check which flux rows did not receive site metadata:

```{r}
flux_model_base %>%
  filter(is.na(IGBP)) %>%
  count(SITE_ID, sort = TRUE) %>%
  head()
```

# ERA5 monthly reanalysis data

The static climate layer (`GlobalClimate.tif`) gives each site one long-term average. ERA5 gives monthly climate values that can be joined to site-month flux rows.

The download workflow is shown below, but the course data folder already includes processed ERA5 rasters:

-   `data/products/era5_ppt_monthly.tif`
-   `data/products/era5_tmean_monthly.tif`

### Download ERA5 data

```{r}
#| eval: false
library(ecmwfr)

era5_years <- 2006:2018

era5_requests <- lapply(era5_years, function(yr) {
  list(
    dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
    product_type = "monthly_averaged_reanalysis",
    variable = c("2m_temperature", "total_precipitation"),
    year = as.character(yr),
    month = sprintf("%02d", 1:12),
    time = "00:00",
    data_format = "netcdf",
    target = paste0("era5_", yr, ".nc")
  )
})

era5_dir <- file.path("data", "products", "era5")
dir.create(era5_dir, showWarnings = FALSE, recursive = TRUE)

wf_request_batch(request_list = era5_requests,
                 workers = 2,
                 retry = 30,
                 user = "ecmwfr",
                 path = era5_dir)
```

::: {.callout-note}
## Why we use the processed rasters below
ERA5 downloads depend on an external account, API key, and queue. For this workshop, we use processed rasters that have already been downloaded and converted to the right units so the data integration workflow is reproducible in class.
:::

### Extract ERA5 values to tower sites

Read the processed monthly ERA5 rasters:

```{r}
era5_ppt_monthly <- rast("data/products/era5_ppt_monthly.tif")
era5_tmean_monthly <- rast("data/products/era5_tmean_monthly.tif")

nlyr(era5_ppt_monthly)
nlyr(era5_tmean_monthly)
```

Name each layer by `YearMon`:

```{r}
names(era5_ppt_monthly) <- as.character(zoo::as.yearmon(names(era5_ppt_monthly)))
names(era5_tmean_monthly) <- as.character(zoo::as.yearmon(names(era5_tmean_monthly)))
```

Transform tower points and extract monthly values:

```{r}
fluxnet_ch4_era5 <- st_transform(fluxnet_ch4_sf, crs = crs(era5_ppt_monthly))

era5_ppt_long <- terra::extract(era5_ppt_monthly, vect(fluxnet_ch4_era5), ID = FALSE) %>%
  mutate(SITE_ID = fluxnet_ch4_era5$SITE_ID) %>%
  pivot_longer(-SITE_ID, names_to = "YearMon", values_to = "era5_p_f") %>%
  mutate(YearMon = zoo::as.yearmon(YearMon))

era5_tmean_long <- terra::extract(era5_tmean_monthly, vect(fluxnet_ch4_era5), ID = FALSE) %>%
  mutate(SITE_ID = fluxnet_ch4_era5$SITE_ID) %>%
  pivot_longer(-SITE_ID, names_to = "YearMon", values_to = "era5_ta_f") %>%
  mutate(YearMon = zoo::as.yearmon(YearMon))

era5_monthly <- era5_ppt_long %>%
  left_join(era5_tmean_long, by = c("SITE_ID", "YearMon"))
```

Check the ERA5 key:

```{r}
find_duplicate_keys(era5_monthly, SITE_ID, YearMon)
range(era5_monthly$YearMon)
```

Join ERA5 onto the modeling table:

```{r}
before_rows <- nrow(flux_model_base)

flux_model_era5 <- flux_model_base %>%
  left_join(era5_monthly, by = c("SITE_ID", "YearMon"))

after_rows <- nrow(flux_model_era5)

before_rows
after_rows
```

# Validate ERA5 against tower measurements

Compare tower-measured precipitation (`P_F`) with ERA5 precipitation:

```{r}
ggplot(flux_model_era5, aes(x = P_F, y = era5_p_f)) +
  geom_point(alpha = 0.45) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(title = "Tower vs. ERA5 monthly precipitation",
       x = "Tower-measured P_F (mm)",
       y = "ERA5 precipitation (mm)")

cor(flux_model_era5$P_F, flux_model_era5$era5_p_f, use = "complete.obs")
```

Compare tower-measured air temperature (`TA_F`) with ERA5 temperature:

```{r}
ggplot(flux_model_era5, aes(x = TA_F, y = era5_ta_f)) +
  geom_point(alpha = 0.45) +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(title = "Tower vs. ERA5 monthly mean temperature",
       x = "Tower-measured TA_F (degrees C)",
       y = "ERA5 temperature (degrees C)")

cor(flux_model_era5$TA_F, flux_model_era5$era5_ta_f, use = "complete.obs")
```

A tower measures its immediate surroundings. ERA5 represents average conditions across a much larger grid cell. Expect closer agreement for temperature than precipitation. This is the kind of check you should do before trusting a reanalysis product as a predictor or substitute for missing observations.

# Build the model-ready table

Now select the response variable, candidate predictors, and useful identifiers. This creates a table that is ready for exploratory modeling.

```{r}
methane_model_data <- flux_model_era5 %>%
  select(SITE_ID, SITE_NAME, YearMon,
         FCH4_F_gC,
         P_F, TA_F, VPD_F,
         NDVI, EVI,
         IGBP, country, continent,
         soil_ph, soil_nitrogen, soil_bulk_density,
         climate_map, climate_tmin, climate_tmax, climate_mat,
         elevation_m,
         era5_p_f, era5_ta_f) %>%
  filter(!is.na(FCH4_F_gC))
```

Inspect the table:

```{r}
glimpse(methane_model_data)
nrow(methane_model_data)
n_distinct(methane_model_data$SITE_ID)
range(methane_model_data$YearMon)
```

# Data quality dashboard

Check missingness:

```{r}
methane_model_data %>%
  count_missing() %>%
  head(25)
```

Check duplicate site-month rows:

```{r}
find_duplicate_keys(methane_model_data, SITE_ID, YearMon)
```

Check which sites have the most complete model records:

```{r}
methane_model_data %>%
  mutate(complete_predictors = complete.cases(P_F, TA_F, VPD_F,
                                              NDVI, EVI,
                                              soil_ph, soil_nitrogen,
                                              climate_map, climate_mat,
                                              elevation_m,
                                              era5_p_f, era5_ta_f)) %>%
  ungroup() %>%
  reframe(.by = SITE_ID,
          total_rows = n(),
          complete_rows = sum(complete_predictors),
          complete_fraction = complete_rows / total_rows) %>%
  arrange(desc(complete_fraction), desc(complete_rows)) %>%
  head(15)
```

Visualize the response variable:

```{r}
ggplot(methane_model_data, aes(x = FCH4_F_gC)) +
  geom_histogram(color = "black", fill = "white") +
  labs(title = "Distribution of monthly methane flux observations",
       x = "FCH4_F_gC",
       y = "Count")
```

# Flux summaries and tower dissimilarity

A useful modeling dataset should help us ask whether towers that are environmentally similar also behave similarly. Here, we summarize methane fluxes by tower and compare those summaries to tower dissimilarity based on integrated predictor data.

First, compute the mean and variability of methane flux for each tower:

```{r}
tower_flux_summary <- methane_model_data %>%
  ungroup() %>%
  reframe(.by = c(SITE_ID, SITE_NAME, IGBP, country, continent),
          n_months = sum(!is.na(FCH4_F_gC)),
          mean_flux = mean(FCH4_F_gC, na.rm = TRUE),
          sd_flux = sd(FCH4_F_gC, na.rm = TRUE),
          flux_cv = sd_flux / abs(mean_flux))
```

Look at the sites with the highest mean methane flux:

```{r}
tower_flux_summary %>%
  arrange(desc(mean_flux)) %>%
  head(10)
```

Look at the sites with the most variable methane flux:

```{r}
tower_flux_summary %>%
  arrange(desc(sd_flux)) %>%
  head(10)
```

Create one predictor summary per tower. Static variables such as soil and elevation are repeated across months, while dynamic variables such as NDVI, precipitation, and temperature are summarized across the available monthly record:

```{r}
tower_predictor_summary <- methane_model_data %>%
  ungroup() %>%
  reframe(.by = SITE_ID,
          mean_p_f = mean(P_F, na.rm = TRUE),
          mean_ta_f = mean(TA_F, na.rm = TRUE),
          mean_vpd_f = mean(VPD_F, na.rm = TRUE),
          mean_ndvi = mean(NDVI, na.rm = TRUE),
          mean_evi = mean(EVI, na.rm = TRUE),
          mean_era5_p_f = mean(era5_p_f, na.rm = TRUE),
          mean_era5_ta_f = mean(era5_ta_f, na.rm = TRUE),
          soil_ph = first(soil_ph),
          soil_nitrogen = first(soil_nitrogen),
          soil_bulk_density = first(soil_bulk_density),
          climate_map = first(climate_map),
          climate_mat = first(climate_mat),
          elevation_m = first(elevation_m))
```

Join flux summaries and predictor summaries:

```{r}
tower_summary <- tower_flux_summary %>%
  left_join(tower_predictor_summary, by = "SITE_ID")
```

Calculate environmental dissimilarity among towers using standardized predictors. Standardizing puts all variables on comparable scales before calculating distance:

```{r}
dissimilarity_vars <- c("mean_p_f", "mean_ta_f", "mean_vpd_f",
                        "mean_ndvi", "mean_evi",
                        "mean_era5_p_f", "mean_era5_ta_f",
                        "soil_ph", "soil_nitrogen", "soil_bulk_density",
                        "climate_map", "climate_mat", "elevation_m")

tower_dissimilarity_input <- tower_summary %>%
  select(SITE_ID, SITE_NAME, IGBP, mean_flux, sd_flux, all_of(dissimilarity_vars)) %>%
  drop_na()

scaled_predictors <- tower_dissimilarity_input %>%
  select(all_of(dissimilarity_vars)) %>%
  scale()

tower_dissimilarity <- dist(scaled_predictors)
```

Find each tower's nearest environmental neighbor:

```{r}
dissimilarity_matrix <- as.matrix(tower_dissimilarity)
diag(dissimilarity_matrix) <- NA

nearest_neighbor_index <- apply(dissimilarity_matrix, 1, which.min)

nearest_tower_summary <- tower_dissimilarity_input %>%
  mutate(nearest_SITE_ID = tower_dissimilarity_input$SITE_ID[nearest_neighbor_index],
         nearest_SITE_NAME = tower_dissimilarity_input$SITE_NAME[nearest_neighbor_index],
         nearest_dissimilarity = dissimilarity_matrix[cbind(seq_len(nrow(dissimilarity_matrix)),
                                                            nearest_neighbor_index)])

nearest_tower_summary %>%
  select(SITE_ID, SITE_NAME, IGBP,
         nearest_SITE_ID, nearest_SITE_NAME, nearest_dissimilarity,
         mean_flux, sd_flux) %>%
  arrange(nearest_dissimilarity) %>%
  head(10)
```

Create a pairwise comparison table. Each row compares two towers:

```{r}
tower_pairs <- as.data.frame(as.table(dissimilarity_matrix)) %>%
  rename(site_1_index = Var1,
         site_2_index = Var2,
         environmental_dissimilarity = Freq) %>%
  mutate(site_1_index = as.integer(site_1_index),
         site_2_index = as.integer(site_2_index)) %>%
  filter(site_1_index < site_2_index) %>%
  mutate(SITE_ID_1 = tower_dissimilarity_input$SITE_ID[site_1_index],
         SITE_ID_2 = tower_dissimilarity_input$SITE_ID[site_2_index],
         mean_flux_difference = abs(tower_dissimilarity_input$mean_flux[site_1_index] -
                                      tower_dissimilarity_input$mean_flux[site_2_index]),
         sd_flux_difference = abs(tower_dissimilarity_input$sd_flux[site_1_index] -
                                    tower_dissimilarity_input$sd_flux[site_2_index]))
```

Do environmentally similar towers have more similar mean methane flux?

```{r}
ggplot(tower_pairs,
       aes(x = environmental_dissimilarity, y = mean_flux_difference)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm") +
  labs(title = "Tower dissimilarity vs. difference in mean methane flux",
       x = "Environmental dissimilarity between towers",
       y = "Absolute difference in mean FCH4_F_gC")

cor(tower_pairs$environmental_dissimilarity,
    tower_pairs$mean_flux_difference,
    use = "complete.obs")
```

Do environmentally similar towers have more similar flux variability?

```{r}
ggplot(tower_pairs,
       aes(x = environmental_dissimilarity, y = sd_flux_difference)) +
  geom_point(alpha = 0.35) +
  geom_smooth(method = "lm") +
  labs(title = "Tower dissimilarity vs. difference in methane flux variability",
       x = "Environmental dissimilarity between towers",
       y = "Absolute difference in SD of FCH4_F_gC")

cor(tower_pairs$environmental_dissimilarity,
    tower_pairs$sd_flux_difference,
    use = "complete.obs")
```

::: {.callout-note}
## Interpreting dissimilarity
This is a screening analysis. If environmentally similar towers have similar methane flux means or variability, that supports using these predictors in a model. If they do not, it suggests important drivers may still be missing, such as water table depth, wetland type, soil carbon, disturbance history, or measurement design.
:::

# Save your work

Save the model-ready table and the ERA5 rasters needed for the spatial projection workshop:

```{r}
#| eval: false
save(methane_model_data, file = "data/products/methane_model_data.RDATA")
writeRaster(era5_ppt_monthly, "data/products/era5_ppt_monthly.tif", overwrite = TRUE)
writeRaster(era5_tmean_monthly, "data/products/era5_tmean_monthly.tif", overwrite = TRUE)
```

# Final project data integration plan

For your final methane model, identify:

1.  Your response variable.
2.  Your spatial extent.
3.  Your temporal resolution.
4.  At least five candidate predictors.
5.  The data source for each predictor.
6.  Whether each predictor is static or dynamic.
7.  The expected units.
8.  The join key or extraction method.
9.  A validation check for each predictor.

## Assignment

Submit a 1-2 page **Final Project Data Integration Plan**. Include:

1.  A short description of the methane flux question you want to answer.
2.  A table of at least five candidate predictors with source, units, spatial resolution, temporal resolution, static/dynamic status, and join key.
3.  One paragraph describing likely missingness, scale mismatch, or join problems.
4.  A short workflow diagram or pseudocode showing how you will compile your modeling table.
5.  One validation check you will use before modeling.
6.  A brief plan for comparing flux mean and variability across sites, including which predictors you would use to calculate tower dissimilarity.
