---
title: "Introduction to Data Integration in R"
editor:
  mode: source
---

Data analysis often requires combining data from multiple sources, such as files, databases, APIs, or web scraping. R is a powerful and flexible tool for data integration, but it can also pose some challenges and pitfalls. In this workshop, you will learn some of the best ways to integrate data from multiple sources in R.

## Getting a Climate Data Store (CDS) API key

Later in this workshop you'll download real historical climate data (ERA5) from the Copernicus Climate Data Store (CDS). This requires a free personal account and API key, separate from anything else in this course, so set it up now — it can take a few minutes for the account to activate.

1. Create a free account at <https://cds.climate.copernicus.eu/>.
2. Once logged in, open your user profile page and copy your **Personal Access Token** — this is your API key.
3. In R, register it once per computer (you'll install the `ecmwfr` package later in this workshop, but you can do it now too):

```{r}
#| eval: false
install.packages("ecmwfr")
library(ecmwfr)
wf_set_key(key = "paste-your-personal-access-token-here")
```

`ecmwfr` stores the key in your system's secure credential manager, so you only need to do this once — every future R session on this computer will use it automatically.

::: {.callout-warning}
## Never save your API key in a script
Type your key directly into the R console (or a chunk you run once and then delete), not into a `.qmd` file you save or push to GitHub. Treat it like a password — if it's ever exposed, regenerate it from your CDS profile page.
:::

#### Workshop Goals:

1. Understand techniques for data Integration.
2. Obtain information from different data sources.
3. Combine tabular and vector data.
4. Combine tabular and raster data.
5. Combine tabular data by an ID and time.

#### Choose the right package

R has many packages that can help you import, merge, and manipulate data from different sources. Some of the most popular and useful ones for tables include readr, dplyr, tidyr, and purrr. These packages are part of the tidyverse, a collection of packages that share a consistent and coherent syntax and philosophy for data analysis. For spatial data, the sf and terra packages are useful.

| Type of Data | Library |
|------:|-----------|
| Tabular | tidyverse |
| Vector | sf |
| Raster | terra |

#### Use pipes and functions

One of the best features of the tidyverse is the pipe operator (%>%), which allows you to chain multiple functions together and pass the output of one function as the input of the next one. This way, you can create a data integration pipeline that is clear and logical, and that avoids intermediate variables and nested functions.

#### Check and validate data

Data integration can introduce errors or inconsistencies in your data, such as duplicates, mismatches, outliers, or invalid values. Therefore, it is important to check and validate your data before and after you integrate it from multiple sources. You can use various tools and techniques to do this, such as summary statistics, data visualization, data profiling, or data quality rules.

In this workshop you will begin to extract information about the FLUXNET CH4 tower sites.

### Install Packages:

```{r}
#| eval: false
install.packages("zoo")
install.packages("rnaturalearth")
install.packages("ecmwfr")
install.packages("lubridate")
```

# Load Libraries:

```{r}
#| output: false
library(sf)
library(terra)
library(tidyverse)
library(ggplot2)
library(rnaturalearth)
library(zoo)
library(ecmwfr)
library(lubridate)
```

# Integrating information from simple features

Import the file FluxNet_Sites_2024.csv and call it fluxnet_sites

```{r}
fluxnet_sites <- read.csv("data/FluxNet_Sites_2024.csv")
```

This dataset includes:

| Column Name | Description |
|---------------------------:|--------------------------------------------|
| SITE_ID | Unique site id |
| SITE_NAME | Site name |
| FLUXNET2015 | License information for the data for the two FLUXNET Products |
| FLUXNET-CH4 | License information for the data for the two FLUXNET Products |
| LOCATION_LAT | Location information |
| LOCATION_LONG | Location information |
| LOCATION_ELEV | Elevation in meters |
| IGBP | Vegetation type |
| MAT | Mean annual temperature in Celsius |
| MAP | Mean annual precipitation in mm |

Subset file to include only sites measuring methane:

```{r}
fluxnet_ch4 <- fluxnet_sites %>% filter(FLUXNET.CH4 == "CC-BY-4.0")
```

Now that you have all of the sites measuring methane, convert fluxnet_ch4 to a sf and call it fluxnet_ch4_sf:

```{r}
fluxnet_ch4_sf <- st_as_sf(x = fluxnet_ch4,
                            coords = c("LOCATION_LONG", "LOCATION_LAT"),
                            crs = 4326)

ggplot(data = fluxnet_ch4_sf) + geom_sf()
```

Check the class and that the geometry is valid:

```{r}
class(fluxnet_ch4_sf)
st_is_valid(fluxnet_ch4_sf)
```

Create a global sf and extract the country into the sf

```{r}
world_sf <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent %in% c("Europe", "Asia", "North America", "South America", "Africa", "Oceania"))

st_is_valid(world_sf)
```

Make the CRS match:

```{r}
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = 4087)

world_sf <- st_transform(world_sf, crs = 4087) %>% st_make_valid()

ggplot() + geom_sf(data = world_sf) + geom_sf(data = fluxnet_ch4_sf)
```

Use st_intersection to extract the country of each tower site. A straight `st_intersection()` can return more or fewer rows than you started with (a point can sit exactly on a shared border and match two polygons, or fall outside all of them), so join back on `SITE_ID` instead of assigning the result directly:

```{r}
country_lookup <- st_intersection(world_sf, fluxnet_ch4_sf) %>%
  st_drop_geometry() %>%
  select(SITE_ID, name) %>%
  distinct(SITE_ID, .keep_all = TRUE)

fluxnet_ch4_sf <- fluxnet_ch4_sf %>%
  left_join(country_lookup, by = "SITE_ID") %>%
  rename(country = name)

fluxnet_ch4_sf$country
```

# Integrating information from rasters

Import the file GlobalSoil_grids.tif :

```{r}
soil <- terra::rast("data/GlobalSoil_grids.tif")
soil
```

Transform fluxnet_ch4_sf to the same CRS as soil:

```{r}
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = crs(soil))
fluxnet_ch4_sf
```

Extract soil information to fluxnet_ch4_sf:

```{r}
fluxnet_ch4_sf$soil_bulk_density <- terra::extract(soil, fluxnet_ch4_sf)$BulkDensity
fluxnet_ch4_sf$soil_bulk_density

fluxnet_ch4_sf$soil_ph <- terra::extract(soil, fluxnet_ch4_sf)$PH
fluxnet_ch4_sf$soil_ph

fluxnet_ch4_sf$soil_nitrogen <- terra::extract(soil, fluxnet_ch4_sf)$Nitrogen
fluxnet_ch4_sf$soil_nitrogen
```

Import the climate information (GlobalClimate.tif) :

```{r}
climate <- terra::rast("data/GlobalClimate.tif")
climate
```

Transform fluxnet_ch4_sf to the same CRS as climate:

```{r}
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = crs(climate))
fluxnet_ch4_sf
```

Look at the data that is available in climate:

```{r}
names(climate)
```

Extract climate information to fluxnet_ch4_sf. Note these are named with a `climate_` prefix rather than reusing `MAT`/`MAP` — `fluxnet_sites` already has site-reported `MAT`/`MAP` columns, and this keeps the raster-extracted values available for comparison instead of silently overwriting them:

```{r}
fluxnet_ch4_sf$climate_map <- terra::extract(climate, fluxnet_ch4_sf)$MAP
fluxnet_ch4_sf$climate_tmin <- terra::extract(climate, fluxnet_ch4_sf)$TMIN
fluxnet_ch4_sf$climate_tmax <- terra::extract(climate, fluxnet_ch4_sf)$TMAX
fluxnet_ch4_sf$climate_mat <- terra::extract(climate, fluxnet_ch4_sf)$MAT

fluxnet_ch4_sf$climate_map
fluxnet_ch4_sf$climate_tmin
fluxnet_ch4_sf$climate_tmax
fluxnet_ch4_sf$climate_mat
```

Import elevation information (Elevation.tif):

```{r}
elevation <- terra::rast("data/Elevation.tif")
elevation
```

Transform fluxnet_ch4_sf to the same CRS as elevation:

```{r}
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = crs(elevation))
```

Look at the data that is available in elevation:

```{r}
names(elevation)
```

Extract elevation information to fluxnet_ch4_sf:

```{r}
fluxnet_ch4_sf$elevation_m <- terra::extract(elevation, fluxnet_ch4_sf)$wc2.1_2.5m_elev
fluxnet_ch4_sf$elevation_m
```

# Joining tables

We can combine columns from two (or more) tables together. This can be achieved using the join family of functions in dplyr. There are different types of joins that will result in different outcomes.

`inner_join()` includes all rows that appear in both the first data frame (x) and the second data frame (y).

`left_join()` returns all rows from x based on matching rows on shared columns in y.
`right_join()` is the companion to `left_join()`, but returns all rows included in y based on matching rows on shared columns in x.

Import APPEEARS file where I requested MODIS NDVI and EVI data for all FLUXNET sites (data/ENV720-MOD13A3-061-results.csv):

```{r}
modis_vi <- read.csv("data/ENV720-MOD13A3-061-results.csv")
names(modis_vi)
```

Subset the modis_vi dataset to include only the columns of interest and rename them:

```{r}
fluxnet_ndvi <- modis_vi %>% select("ID",
                                     "Date",
                                     "MOD13A3_061__1_km_monthly_EVI", "MOD13A3_061__1_km_monthly_NDVI", "MOD13A3_061__1_km_monthly_VI_Quality") %>%
  rename(SITE_ID = ID,
         EVI = MOD13A3_061__1_km_monthly_EVI,
         NDVI = MOD13A3_061__1_km_monthly_NDVI,
         QAQC = MOD13A3_061__1_km_monthly_VI_Quality) %>%
  filter(QAQC > 0)

names(fluxnet_ndvi)
```

Make the fluxnet_ch4_sf vector a dataframe:

```{r}
fluxnet_ch4_df <- as.data.frame(fluxnet_ch4_sf)
```

Identify the column that you should use to join the datasets fluxnet_ch4_df and fluxnet_ndvi:

```{r}
#| output: false
fluxnet_ch4_df$SITE_ID
fluxnet_ndvi$SITE_ID
```

Use left_join because you want to keep all the data from fluxnet_ch4_df and only the sites in fluxnet_ndvi that match the sites in fluxnet_ch4_df:

```{r}
fluxnet_ch4_final <- fluxnet_ch4_df %>% left_join(fluxnet_ndvi, by = "SITE_ID")
```

Check to make sure the list of sites matches:

```{r}
length(unique(fluxnet_ch4_df$SITE_ID))
length(unique(fluxnet_ch4_final$SITE_ID))
```

Import the monthly FLUX data:

```{r}
load("data/FLUXNET_FLUXES.RDATA")

# The saved object is called FLUXNET.flux; alias it to match this course's naming convention
fluxnet_flux <- FLUXNET.flux
```

Look at the flux file "fluxnet_flux":

```{r}
summary(fluxnet_flux)
names(fluxnet_flux)
```

```{r}
#| output: false
# Make SITE ID:
fluxnet_flux$SITE_ID <- fluxnet_flux$SITE
fluxnet_flux$SITE_ID

# This is a good column to join based on:
fluxnet_flux$YearMon
```

In your site file, convert the date to Year-Month and use this column to join it with the flux file "fluxnet_flux":

```{r}
names(fluxnet_ch4_final)
summary(fluxnet_ch4_final$Date)
```

```{r}
#| include: false
#| output: false
fluxnet_ch4_final$Date
```

```{r}
# Format as a Date:
fluxnet_ch4_final$date_formatted <- fluxnet_ch4_final$Date %>% as.Date(format = "%Y-%m-%d")

# Format as a Yearmon:
fluxnet_ch4_final$YearMon <- fluxnet_ch4_final$date_formatted %>% zoo::as.yearmon("%m-%Y")

class(fluxnet_ch4_final$YearMon)
```

Join the two files:

```{r}
fluxes_month <- fluxnet_flux %>% left_join(fluxnet_ch4_final, by = c("YearMon", "SITE_ID"))
```

Save your file:

```{r}
#| eval: false
save(fluxes_month, file = "data/products/Monthly_Fluxes.RDATA")
```

# Validating tower measurements with ERA5 reanalysis climate data

The static climate layer you extracted earlier (GlobalClimate.tif) gives each site a single long-term average — useful for describing a site, but not for matching a specific month. `fluxnet_flux` already includes tower-measured monthly precipitation (P_F) and temperature (TA_F), recorded by instruments at each site. Here we'll download ERA5, a global atmospheric reanalysis produced by the European Centre for Medium-Range Weather Forecasts (ECMWF), and compare its monthly values against what the towers actually measured. ERA5 is genuinely global and covers specific months back to 1940, which is why we're using it here instead of the climate normals from earlier workshops — and the same download will double as the input for your spatial projection workshop.

Register your CDS key for this R session, if you haven't already (see the setup instructions at the top of this workshop):

```{r}
#| eval: false
wf_set_key(key = "paste-your-personal-access-token-here")
```

### Download monthly ERA5 data

We need monthly precipitation and temperature for every year covered by `fluxnet_flux`:

```{r}
range(fluxnet_flux$YearMon)
```

Build one request per year (global extent, both variables), then submit them together with `wf_request_batch()`, which runs several requests at once instead of waiting for each one in turn:

```{r}
era5_years <- 2006:2018

era5_requests <- lapply(era5_years, function(yr) {
  list(
    dataset_short_name = "reanalysis-era5-single-levels-monthly-means",
    product_type = "monthly_averaged_reanalysis",
    variable = c("2m_temperature", "total_precipitation"),
    year = as.character(yr),
    month = sprintf("%02d", 1:12),
    time = "00:00",
    data_format = "netcdf",
    target = paste0("era5_", yr, ".nc")
  )
})
```

```{r}
#| eval: false
era5_dir <- file.path("data", "products", "era5")
dir.create(era5_dir, showWarnings = FALSE, recursive = TRUE)

wf_request_batch(request_list = era5_requests, workers = 2, retry = 30, user = "ecmwfr", path = era5_dir)
```

::: {.callout-note}
## This step takes a few minutes
Each request is processed by ECMWF's queue rather than downloaded instantly — the whole batch (13 years, global, 2 variables) typically finishes in a few minutes, but can take longer if the CDS service is busy. You only need to run it once; the files are saved to `data/products/era5/`. Don't raise `workers` or lower `retry` to try to speed this up — polling too aggressively gets your requests throttled by CDS ("check your retry rate") rather than completed faster.
:::

Each year downloads as a zip containing one NetCDF file per variable. Unzip them into per-year folders so files with the same name across years don't collide:

```{r}
era5_dir <- file.path("data", "products", "era5")
era5_zips <- list.files(era5_dir, pattern = "[.]zip$", full.names = TRUE)

for (z in era5_zips) {
  unzip(z, exdir = tools::file_path_sans_ext(z))
}
```

Read each variable's files across all years into one SpatRaster stack, in chronological order:

```{r}
era5_ppt_files <- list.files(era5_dir, pattern = "avgad[.]nc$", recursive = TRUE, full.names = TRUE) %>% sort()
era5_tmean_files <- list.files(era5_dir, pattern = "avgua[.]nc$", recursive = TRUE, full.names = TRUE) %>% sort()

era5_ppt_raw <- rast(era5_ppt_files)
era5_tmean_raw <- rast(era5_tmean_files)

nlyr(era5_ppt_raw)
```

ERA5's global grid runs longitude 0-360 degrees, but our site coordinates use the standard -180 to 180 degree convention. `terra::rotate()` converts it so the two line up:

```{r}
era5_ppt_raw <- terra::rotate(era5_ppt_raw)
era5_tmean_raw <- terra::rotate(era5_tmean_raw)
```

### Convert units to match P_F and TA_F

`total_precipitation` in this product is a mean *daily* rate in meters, not a monthly total, so we multiply by 1000 (meters to mm) and by the number of days in that layer's month:

```{r}
era5_ppt_mm <- era5_ppt_raw * 1000
era5_ppt_monthly <- era5_ppt_mm * lubridate::days_in_month(terra::time(era5_ppt_raw))
```

`2m_temperature` is already a monthly mean, in Kelvin — just convert to Celsius:

```{r}
era5_tmean_c <- era5_tmean_raw - 273.15
```

### Extract monthly values at each tower site

Name each layer with its year-month so we can pivot the extracted values into a long, joinable format:

```{r}
names(era5_ppt_monthly) <- as.character(zoo::as.yearmon(terra::time(era5_ppt_raw)))
names(era5_tmean_c) <- as.character(zoo::as.yearmon(terra::time(era5_tmean_raw)))

fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = crs(era5_ppt_monthly))

era5_ppt_long <- terra::extract(era5_ppt_monthly, fluxnet_ch4_sf, ID = FALSE) %>%
  mutate(SITE_ID = fluxnet_ch4_sf$SITE_ID) %>%
  pivot_longer(-SITE_ID, names_to = "YearMon", values_to = "era5_p_f") %>%
  mutate(YearMon = zoo::as.yearmon(YearMon))

era5_tmean_long <- terra::extract(era5_tmean_c, fluxnet_ch4_sf, ID = FALSE) %>%
  mutate(SITE_ID = fluxnet_ch4_sf$SITE_ID) %>%
  pivot_longer(-SITE_ID, names_to = "YearMon", values_to = "era5_ta_f") %>%
  mutate(YearMon = zoo::as.yearmon(YearMon))

era5_monthly <- era5_ppt_long %>% left_join(era5_tmean_long, by = c("SITE_ID", "YearMon"))
```

Join ERA5 onto the flux data, the same way you joined NDVI/EVI earlier:

```{r}
fluxes_month_era5 <- fluxes_month %>% left_join(era5_monthly, by = c("SITE_ID", "YearMon"))
```

### How well does ERA5 match the tower measurements?

```{r}
ggplot(fluxes_month_era5, aes(x = P_F, y = era5_p_f)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(title = "Tower vs. ERA5 monthly precipitation",
       x = "Tower-measured P_F (mm)",
       y = "ERA5 precipitation (mm)")

cor(fluxes_month_era5$P_F, fluxes_month_era5$era5_p_f, use = "complete.obs")
```

```{r}
ggplot(fluxes_month_era5, aes(x = TA_F, y = era5_ta_f)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, col = "red") +
  labs(title = "Tower vs. ERA5 monthly mean temperature",
       x = "Tower-measured TA_F (degrees C)",
       y = "ERA5 temperature (degrees C)")

cor(fluxes_month_era5$TA_F, fluxes_month_era5$era5_ta_f, use = "complete.obs")
```

A tower sits at one exact point and measures its immediate surroundings; ERA5 is a reanalysis model output at ~0.25 degree (~28 km) resolution, so each value represents the average condition across a fairly large grid cell. Expect closer agreement for temperature (which varies smoothly over that distance) than for precipitation (which is often much more localized) — this is exactly the kind of gap to check for before trusting a reanalysis product as a stand-in for missing tower data.

### Save your work

Save the ERA5-enriched dataset, and the two raster stacks so you don't need to re-download them for the spatial projection workshop:

```{r}
#| eval: false
save(fluxes_month_era5, file = "data/products/fluxes_month_era5.RDATA")
writeRaster(era5_ppt_monthly, "data/products/era5_ppt_monthly.tif", overwrite = TRUE)
writeRaster(era5_tmean_c, "data/products/era5_tmean_monthly.tif", overwrite = TRUE)
```

You are now prepared to take data from different sources to build a file to explore patterns in methane infrastructure. Take note of the difference between joining site based static data versus site based data that changes over time. In your next in class assessment, you will build a file with all of the data you want to use to develop your methane model. To prepare for this, find and obtain your data sources.
