---
title: "Random Forest Models: Predicting Methane Flux"
editor:
  mode: source
---

## From integrated data to prediction

In Data Integration, we built the kind of table needed for methane flux modeling: one row per site-month, with methane flux as the response and climate, vegetation, and ecosystem variables as candidate predictors.

In this workshop, we will ask:

> Can we use climate, vegetation, and ecosystem predictors to estimate monthly methane flux, and how do we know whether the model is useful?

By the end of this workshop, you will be able to:

1.  Identify a response variable and candidate predictors.
2.  Prepare a model-ready table for random forest regression.
3.  Compare candidate random forest models.
4.  Evaluate train and test performance using RMSE, MAE, R-squared, and bias.
5.  Interpret variable importance without treating it as causation.
6.  Compare random row splits with leave-one-site-out assessment.
7.  Build a final model by selecting the best predictor from each category.
8.  Save a final model object for sensitivity analysis and spatial prediction.

### Install packages

```{r}
#| eval: false
install.packages("randomForest")
install.packages("tidyverse")
install.packages("GGally")
```

### Load libraries

```{r}
#| output: false
library(randomForest)
library(tidyverse)
library(GGally)
```

# Random forests

Random forests are ensembles of decision trees. Each tree is fit to a bootstrap sample of the data, and each split considers only a random subset of predictors. The final prediction is the average prediction across many trees.

Random forests are useful for environmental data because they can model nonlinear relationships and interactions without requiring us to specify those relationships in advance. They are still models, though. They can overfit, they can inherit bias from the training data, and they cannot reliably predict conditions not represented in the data.

# Load the modeling data

This workshop uses a model-ready version of the methane flux dataset. The same structure is what you are building toward in your final project.

```{r}
load("data/RANDOMFOREST_DATASET.RDATA")
```

Inspect the data:

```{r}
glimpse(fluxnet)
names(fluxnet)
```

The response variable is monthly methane flux:

```{r}
summary(fluxnet$FCH4_F_gC)
```

# Modeling checklist

Before fitting any model, identify:

-   response variable: `FCH4_F_gC`;
-   climate predictors: `P_F`, `TA_F`, `VPD_F`;
-   vegetation predictors: `NDVI`, `EVI`;
-   ecosystem predictor: `IGBP`;
-   grouping variable for site-level assessment: `SITE`.

Prepare the modeling table:

```{r}
model_data <- fluxnet %>%
  ungroup() %>%
  select(SITE, YearMon, FCH4_F_gC,
         P_F, TA_F, VPD_F,
         NDVI, EVI,
         IGBP) %>%
  drop_na()
```

Check the table:

```{r}
nrow(model_data)
n_distinct(model_data$SITE)
summary(model_data)
```

Some ecosystem classes occur at only one site. That causes problems for leave-one-site-out prediction: if the held-out site is the only example of an ecosystem class, the training set has never seen that class. Collapse rare classes into `Other`.

```{r}
common_igbp <- model_data %>%
  distinct(SITE, IGBP) %>%
  count(IGBP, name = "n_sites") %>%
  filter(n_sites >= 3) %>%
  pull(IGBP)

model_data <- model_data %>%
  mutate(IGBP_model = if_else(IGBP %in% common_igbp, IGBP, "Other"),
         IGBP_model = as.factor(IGBP_model))
```

Check the collapsed ecosystem classes:

```{r}
model_data %>%
  distinct(SITE, IGBP_model) %>%
  count(IGBP_model, name = "n_sites") %>%
  arrange(desc(n_sites))
```

# Explore the predictors

Visualize key variables before modeling:

```{r}
#| message: false
ggpairs(model_data,
        columns = c("FCH4_F_gC", "P_F", "TA_F", "VPD_F", "NDVI", "EVI"))
```

Look at methane flux by ecosystem type:

```{r}
ggplot(model_data, aes(x = IGBP_model, y = FCH4_F_gC)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Monthly methane flux by ecosystem class",
       x = "IGBP",
       y = "FCH4_F_gC")
```

# Evaluation metrics

We will evaluate models using four metrics:

-   RMSE: root mean squared error; lower is better.
-   MAE: mean absolute error; lower is better.
-   R-squared: squared correlation between observed and predicted; higher is better.
-   Bias: mean predicted minus observed; closer to zero is better.

```{r}
model_metrics <- function(observed, predicted) {
  tibble(
    rmse = sqrt(mean((observed - predicted)^2, na.rm = TRUE)),
    mae = mean(abs(observed - predicted), na.rm = TRUE),
    r2 = cor(observed, predicted, use = "complete.obs")^2,
    bias = mean(predicted - observed, na.rm = TRUE)
  )
}
```

# Train/test split

A random row split asks: can the model predict new months from a network of sites that are mostly represented in the training data?

```{r}
set.seed(111)

model_data <- model_data %>%
  mutate(row_id = row_number())

train <- model_data %>%
  sample_frac(0.80)

test <- model_data %>%
  anti_join(train %>% select(row_id), by = "row_id")

nrow(train)
nrow(test)
```

# Candidate models

Fit three candidate models:

1.  Climate-only model.
2.  Climate + vegetation model.
3.  Climate + vegetation + ecosystem model.

```{r}
candidate_formulas <- list(
  climate = FCH4_F_gC ~ P_F + TA_F + VPD_F,
  climate_vegetation = FCH4_F_gC ~ P_F + TA_F + VPD_F + NDVI + EVI,
  climate_vegetation_ecosystem = FCH4_F_gC ~ P_F + TA_F + VPD_F + NDVI + EVI + IGBP_model
)
```

Fit each model:

```{r}
set.seed(111)

candidate_models <- lapply(candidate_formulas, function(formula) {
  randomForest(formula,
               data = train,
               keep.forest = TRUE,
               importance = TRUE,
               mtry = 2,
               ntree = 300)
})
```

Compare train and test performance:

```{r}
candidate_metrics <- imap_dfr(candidate_models, function(model, model_name) {
  train_pred <- predict(model, train)
  test_pred <- predict(model, test)

  bind_rows(
    model_metrics(train$FCH4_F_gC, train_pred) %>%
      mutate(model = model_name, dataset = "train"),
    model_metrics(test$FCH4_F_gC, test_pred) %>%
      mutate(model = model_name, dataset = "test")
  )
})

candidate_metrics
```

Visualize model performance:

```{r}
candidate_metrics %>%
  pivot_longer(cols = c(rmse, mae, r2, bias),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = model, y = value, fill = dataset)) +
  geom_col(position = "dodge") +
  facet_wrap(~ metric, scales = "free_y") +
  coord_flip() +
  labs(title = "Train and test performance for candidate random forest models",
       x = "Model",
       y = "Metric value")
```

Select the best candidate model by test RMSE:

```{r}
best_candidate_name <- candidate_metrics %>%
  filter(dataset == "test") %>%
  arrange(rmse) %>%
  slice(1) %>%
  pull(model)

best_candidate_name
```

# Diagnose the best candidate model

```{r}
best_candidate_model <- candidate_models[[best_candidate_name]]

test$pred_fch4 <- predict(best_candidate_model, test)
```

Observed vs. predicted:

```{r}
ggplot(test, aes(x = FCH4_F_gC, y = pred_fch4)) +
  geom_point(alpha = 0.65) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Observed vs. predicted methane flux",
       x = "Observed FCH4_F_gC",
       y = "Predicted FCH4_F_gC")
```

Residuals:

```{r}
test <- test %>%
  mutate(residual = FCH4_F_gC - pred_fch4)

ggplot(test, aes(x = pred_fch4, y = residual)) +
  geom_point(alpha = 0.65) +
  geom_hline(yintercept = 0, color = "red") +
  labs(title = "Residuals vs. predicted methane flux",
       x = "Predicted FCH4_F_gC",
       y = "Residual")

ggplot(test, aes(x = residual)) +
  geom_histogram(color = "black", fill = "white") +
  labs(title = "Residual distribution",
       x = "Residual",
       y = "Count")
```

Variable importance:

```{r}
importance(best_candidate_model)
varImpPlot(best_candidate_model)
```

::: {.callout-note}
## Importance is not causation
Variable importance tells us which predictors helped this model make predictions, given this dataset and the other predictors in the model. It does not prove that a predictor causes methane flux to change.
:::

# Leave-one-site-out assessment

A random row split is useful, but it can be optimistic because rows from the same tower may appear in both training and testing. A harder question is:

> Can the model predict methane flux at a tower site it has never seen?

Leave-one-site-out assessment fits one model per site. Each time, one site is held out for testing and all other sites are used for training.

To keep the workshop renderable, we use fewer trees for this assessment.

```{r}
leave_one_site_out <- function(formula, data, ntree = 150, mtry = 2) {
  sites <- sort(unique(data$SITE))

  map_dfr(sites, function(site_id) {
    train_site <- data %>% filter(SITE != site_id)
    test_site <- data %>% filter(SITE == site_id)

    fit <- randomForest(formula,
                        data = train_site,
                        keep.forest = TRUE,
                        importance = FALSE,
                        mtry = mtry,
                        ntree = ntree)

    pred <- predict(fit, test_site)

    model_metrics(test_site$FCH4_F_gC, pred) %>%
      mutate(SITE = site_id,
             n_test = nrow(test_site))
  })
}
```

Run leave-one-site-out assessment for the best candidate model:

```{r}
set.seed(111)

loso_metrics <- leave_one_site_out(candidate_formulas[[best_candidate_name]],
                                   model_data,
                                   ntree = 150,
                                   mtry = 2)

loso_metrics %>%
  arrange(rmse) %>%
  head(10)
```

Which sites are hardest to predict?

```{r}
loso_metrics %>%
  arrange(desc(rmse)) %>%
  head(10)
```

Summarize leave-one-site-out performance across sites:

```{r}
loso_metrics %>%
  summarize(mean_rmse = mean(rmse, na.rm = TRUE),
            median_rmse = median(rmse, na.rm = TRUE),
            mean_mae = mean(mae, na.rm = TRUE),
            median_r2 = median(r2, na.rm = TRUE),
            mean_bias = mean(bias, na.rm = TRUE))
```

Visualize site-level RMSE:

```{r}
ggplot(loso_metrics,
       aes(x = reorder(SITE, rmse), y = rmse)) +
  geom_col() +
  coord_flip() +
  labs(title = "Leave-one-site-out RMSE by held-out site",
       x = "Held-out site",
       y = "RMSE")
```

::: {.callout-note}
## Two meanings of prediction
A random row split evaluates prediction for new observations from a familiar network of sites. Leave-one-site-out evaluates prediction for a new site. For final projects, be clear about which kind of prediction you want your model to support.
:::

# Select the best predictor from each category

For final model development, we will choose one strong predictor from each category:

-   climate: `P_F`, `TA_F`, or `VPD_F`;
-   vegetation: `NDVI` or `EVI`;
-   ecosystem: `IGBP_model`.

We will screen individual predictors using test RMSE. This is a simple strategy, not the only valid one.

```{r}
screen_predictor <- function(predictor, category) {
  formula <- reformulate(predictor, response = "FCH4_F_gC")

  set.seed(111)
  fit <- randomForest(formula,
                      data = train,
                      keep.forest = TRUE,
                      importance = TRUE,
                      mtry = 1,
                      ntree = 300)

  pred <- predict(fit, test)

  model_metrics(test$FCH4_F_gC, pred) %>%
    mutate(category = category,
           predictor = predictor)
}

predictor_screen <- bind_rows(
  map_dfr(c("P_F", "TA_F", "VPD_F"), screen_predictor, category = "climate"),
  map_dfr(c("NDVI", "EVI"), screen_predictor, category = "vegetation"),
  screen_predictor("IGBP_model", category = "ecosystem")
)

predictor_screen
```

Select the best predictor from each category:

```{r}
best_predictors <- predictor_screen %>%
  group_by(category) %>%
  arrange(rmse, .by_group = TRUE) %>%
  slice(1) %>%
  ungroup()

best_predictors
```

Build the final formula:

```{r}
final_predictors <- best_predictors$predictor
final_formula <- reformulate(final_predictors, response = "FCH4_F_gC")
final_formula
```

Fit the final model:

```{r}
set.seed(111)

fch4_rf <- randomForest(final_formula,
                        data = train,
                        keep.forest = TRUE,
                        importance = TRUE,
                        mtry = max(1, floor(length(final_predictors) / 3)),
                        ntree = 500,
                        keep.inbag = TRUE)

fch4_rf
```

Evaluate the final model on train and test data:

```{r}
train$pred_final <- predict(fch4_rf, train)
test$pred_final <- predict(fch4_rf, test)

final_metrics <- bind_rows(
  model_metrics(train$FCH4_F_gC, train$pred_final) %>%
    mutate(dataset = "train"),
  model_metrics(test$FCH4_F_gC, test$pred_final) %>%
    mutate(dataset = "test")
)

final_metrics
```

Plot final model predictions:

```{r}
ggplot(test, aes(x = FCH4_F_gC, y = pred_final)) +
  geom_point(alpha = 0.65) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(title = "Final selected random forest model",
       x = "Observed FCH4_F_gC",
       y = "Predicted FCH4_F_gC")
```

Inspect final model importance:

```{r}
importance(fch4_rf)
varImpPlot(fch4_rf)
```

# Save your final model

Save the final model and associated objects for sensitivity analysis and spatial projection. The object is named `fch4_rf` so later workshops can use it.

```{r}
#| eval: false
save(fch4_rf, train, test, final_formula, final_predictors,
     final_metrics, predictor_screen, loso_metrics,
     file = "data/products/random_forest_model.RDATA")
```

# Assignment

Build and compare at least three random forest models for methane flux:

1.  A climate-only model.
2.  A climate + vegetation model.
3.  A final model that selects the best predictor from each category.

Include:

1.  The model formulas.
2.  Train/test performance metrics.
3.  Leave-one-site-out performance for your preferred model.
4.  Observed vs. predicted plots.
5.  Variable importance for your final model.
6.  One paragraph explaining which model you would use and why.
7.  One paragraph explaining whether the model should be trusted for new tower sites.
