---
title: "Sensitivity Analysis"
editor:
  mode: source
---

A sensitivity analysis is a technique used to understand how changes in the inputs of a mathematical model or system affect the output. In the context of a model, such as a simulation or a statistical model, sensitivity analysis helps to identify which input parameters have the most significant influence on the results.

To begin load the model and dataset we will use for this workshop:

```{r}
load("data/SensitivityProducts.RDATA")

# The saved model object is called FCH4_F_gC.rf; alias it to match this course's naming convention
fch4_rf <- FCH4_F_gC.rf
```

We will use the following packages:

```{r}
#| output: false
library(randomForest)
library(tidyverse)
library(gtools)
library(ggplot2)
```

In this workshop, we will prepare a sensitivity analysis for the model fch4_rf. Take a look at the model:

```{r}
fch4_rf
```

The model includes monthly precipitation (P_F), mean temperature (TA_F), and a binary indicator for upland (1 = upland ecosystem and 0 = aquatic ecosystem).

Explore the conditions present within the dataset (fluxnet.new) used to build the model. First subset only the variables used in the final model:

```{r}
model_vars <- fluxnet.new %>% as.data.frame() %>% select(P_F, TA_F, Upland)
```

Next, we will summarize the conditions within the dataset by quantiles (0.25, 0.5, 0.75): note that your categorical variables need to be listed in .by=c().

```{r}
model_vars_lower <- model_vars %>% reframe(.by = Upland, P_F = quantile(P_F, 0.25),
                                            TA_F = quantile(TA_F, 0.25),
                                            Quantile = as.factor(0.25)) %>% as.data.frame()

model_vars_median <- model_vars %>% reframe(.by = Upland, P_F = quantile(P_F, 0.5),
                                             TA_F = quantile(TA_F, 0.5),
                                             Quantile = as.factor(0.5)) %>% as.data.frame()

model_vars_upper <- model_vars %>% reframe(.by = Upland, P_F = quantile(P_F, 0.75),
                                            TA_F = quantile(TA_F, 0.75),
                                            Quantile = as.factor(0.75)) %>% as.data.frame()
```

Combine the individual summaries into one dataframe using `smartbind()` (from the gtools package), which row-binds data frames even when their columns don't line up exactly, filling in NA for anything missing:

```{r}
condition_summary <- smartbind(model_vars_lower, model_vars_median, model_vars_upper)
condition_summary
```

Now, choose the variable you want to explore: TA_F

Look at the conditions present within the dataset for TA_F

```{r}
summary(fluxnet.new$TA_F)
```

Look at the range in values for TA_F:

```{r}
range(fluxnet.new$TA_F)
```

Access each individual range value:

```{r}
range(fluxnet.new$TA_F)[1]

range(fluxnet.new$TA_F)[2]
```

Use the range to generate a sequence of values going from the highest to lowest:

```{r}
ta_f_seq <- seq(range(fluxnet.new$TA_F)[1], range(fluxnet.new$TA_F)[2], by = 5)
```

Create a dataframe:

```{r}
ta_f_seq_df <- data.frame(TA_F = ta_f_seq)
```

Next we will take condition_summary, remove the values for TA_F and replace it with the generated range using `cross_join()`, which returns every combination of rows between the two data frames (the Cartesian product):

```{r}
ta_f_one_at_a_time <- condition_summary %>% select(-TA_F) %>% cross_join(ta_f_seq_df)
```

In this dataframe, all other variables in the model are held at their quantile values and only temperature varies across the range of values observed. This approach is "one-at-a-time". We choose one variable to vary and hold all others at a specific value. Often the mean is chosen but using the 0.25, 0.5, and 0.75 quantile helps to visualize what the model would predict across conditions.

Next, use the `predict()` function to predict values into the dataframe (ta_f_one_at_a_time) with the simulated conditions:

```{r}
ta_f_one_at_a_time$PRED <- predict(fch4_rf, newdata = ta_f_one_at_a_time)
ta_f_one_at_a_time$PRED
```

Let's look at the predictions.

```{r}
ggplot() + geom_point(data = ta_f_one_at_a_time, aes(x = TA_F, y = PRED))
```

Use geom_smooth to visualize the general relationship:

```{r}
ggplot() + geom_smooth(data = ta_f_one_at_a_time, aes(x = TA_F, y = PRED))
```

Look at the predictions by Quantile and by the Upland indicator:

```{r}
ggplot() + geom_line(data = ta_f_one_at_a_time, aes(x = TA_F, y = PRED, color = Quantile, group = interaction(Quantile, Upland), linetype = Upland))
```

Now explore the next variable: P_F

```{r}
summary(fluxnet.new$P_F)

range(fluxnet.new$P_F)

range(fluxnet.new$P_F)[1]

range(fluxnet.new$P_F)[2]

p_f_seq <- seq(range(fluxnet.new$P_F)[1], range(fluxnet.new$P_F)[2], by = 100)

p_f_seq_df <- data.frame(P_F = p_f_seq)

p_f_one_at_a_time <- condition_summary %>% select(-P_F) %>% cross_join(p_f_seq_df)

p_f_one_at_a_time$PRED <- predict(fch4_rf, newdata = p_f_one_at_a_time)

ggplot() + geom_point(data = p_f_one_at_a_time, aes(x = P_F, y = PRED))

ggplot() + geom_smooth(data = p_f_one_at_a_time, aes(x = P_F, y = PRED))

ggplot() + geom_line(data = p_f_one_at_a_time, aes(x = P_F, y = PRED, color = Quantile, group = interaction(Quantile, Upland), linetype = Upland))
```

Use this same workflow to explore the sensitivity of your groups project's model. Please provide a presentation explaining how you fit your model (mtry?, ntree?) and how variables where selected (forward versus backward selection). Describe your final model results (variables in the final model, their importance, the %Var, observed versus predicted for testing and training data), including a correlation plot, variance importance plots, and sensitivity analyses. This report will develop into the methods and results portion of your final project.
