---
title: "Introduction to Simple Features in R"
editor:
  mode: source
---

### Install libraries for this workshop

```{r}
#| eval: false
install.packages("sf")
install.packages("rnaturalearth")
install.packages("devtools")
devtools::install_github("ropensci/rnaturalearthhires")
devtools::install_github("valentinitnelav/plotbiomes")
```

### Load the required libraries for this workshop

```{r}
#| output: false
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)
library(tidyverse)
library(ggplot2)
```

# Goals

The goals of this workshop are to:

1. Become familiar with simple features
2. Master simple feature manipulation
3. Visualize simple features

# Data

Import FluxNet_Sites_2024.csv. This table was created from the FLUXNET site list found at <https://fluxnet.org/sites/site-list-and-pages/?view=table>.

```{r}
fluxnet_sites <- read.csv("data/FluxNet_Sites_2024.csv")
```

This dataset includes:

| Column Name | Description |
|---------------------------:|--------------------------------------------|
| SITE_ID | Unique site id |
| SITE_NAME | Site name |
| FLUXNET2015 | License information for the data for the two FLUXNET Products |
| FLUXNET-CH4 | License information for the data for the two FLUXNET Products |
| LOCATION_LAT | Location information |
| LOCATION_LONG | Location information |
| LOCATION_ELEV | Elevation in meters |
| IGBP | Vegetation type |
| MAT | Mean annual temperature in Celsius |
| MAP | Mean annual precipitation in mm |

Take a look at the file:

```{r}
#| eval: false
View(fluxnet_sites)
```

Look at the tower site locations:

```{r}
fluxnet_sites %>% ggplot() + geom_point(aes(x = LOCATION_LONG, y = LOCATION_LAT))
```

Look at the elevation, mean annual temperature, and mean annual precipitation for the tower site locations:

```{r}
fluxnet_sites %>% ggplot(aes(x = LOCATION_ELEV)) +
  geom_histogram(color = "black", fill = "white")

fluxnet_sites %>% ggplot(aes(x = MAT)) +
  geom_histogram(color = "black", fill = "white")

fluxnet_sites %>% ggplot(aes(x = MAP)) +
  geom_histogram(color = "black", fill = "white")
```

We are interested in exploring the sites with methane data. Let's subset by FLUXNET-CH4 (read into R as the column `FLUXNET.CH4`, since R replaces the hyphen with a dot).

```{r}
fluxnet_ch4 <- fluxnet_sites %>% filter(FLUXNET.CH4 != "")
```

```{r}
#| eval: false
View(fluxnet_ch4)
```

This object is currently a dataframe.

```{r}
class(fluxnet_ch4)
```

Let's make it a simple feature using `st_as_sf()`.

```{r}
fluxnet_ch4_sf <- st_as_sf(x = fluxnet_ch4,
                            coords = c("LOCATION_LONG", "LOCATION_LAT"),
                            crs = 4326)

ggplot(data = fluxnet_ch4_sf) + geom_sf()
```

Check the class:

```{r}
class(fluxnet_ch4_sf)
```

Simple features describe how objects in the real world can be represented in computers. They have a geometry describing where on earth the feature is located, and they have attributes, which describe other properties about the feature.

Look at the information about the geometry:

```{r}
fluxnet_ch4_sf$geometry
```

If we print the first three features, we see their attribute values and an abridged version of the geometry.

```{r}
print(fluxnet_ch4_sf, n = 3)
```

# Geometrical Operations

There are many geometrical operations that can be used to achieve simple feature manipulation.

```{r}
methods(class = "sf")
```

Below we will explore a few methods:

`st_is_valid()` and `st_is_simple()` return a boolean indicating whether a geometry is valid or simple.

```{r}
st_is_valid(fluxnet_ch4_sf)
```

If there were any issues with the geometry you can use `st_make_valid()` to fix it. Since this is a point file there will be no issues with the geometry.

### Change the CRS to a projected EPSG with st_transform.

<a href="https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf"> Coordinate reference systems (CRS) </a> are like measurement units for coordinates: they specify which location on Earth a particular coordinate pair refers to. Simple features have two attributes to store a CRS: epsg and proj4string. This implies that all geometries in a geometry list-column must have the same CRS. Both may be NA, e.g. in case the CRS is unknown, or when we work with local coordinate systems (e.g. inside a building, a body, or an abstract space).

proj4string is a generic, string-based description of a CRS, understood by the PROJ library. It defines projection types and (often) defines parameter values for particular projections, and hence can cover an infinite amount of different projections. This library (also used by GDAL) provides functions to convert or transform between different CRS. epsg is the integer ID for a particular, known CRS that can be resolved into a proj4string. Some proj4string values can be resolved back into their corresponding epsg ID, but this does not always work.

The importance of having epsg values stored with data besides proj4string values is that the epsg refers to a particular, well-known CRS, whose parameters may change (improve) over time; fixing only the proj4string may remove the possibility to benefit from such improvements, and limit some of the provenance of datasets, but may help reproducibility. Coordinate reference system transformations can be carried out using st_transform.

```{r}
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = 4087)

fluxnet_ch4_sf
```

Now the projected CRS is WGS 84 / World Equidistant Cylindrical.

`st_distance()` returns a dense numeric matrix with distances between geometries:

```{r}
fluxnet_ch4_sf$mean_distance_km <- st_distance(fluxnet_ch4_sf, fluxnet_ch4_sf) %>% rowMeans() / 1000

fluxnet_ch4_sf %>% ggplot() + geom_sf(aes(col = mean_distance_km))

fluxnet_ch4_sf %>% ggplot(aes(x = mean_distance_km)) +
  geom_histogram(color = "black", fill = "white")
```

# Getting country and state boundaries with rnaturalearth

rnaturalearth provides world map data at several resolutions, including country boundaries with a `continent` attribute you can filter or group by. Its companion package, rnaturalearthhires, adds higher-resolution boundaries down to the state/province level for individual countries. Together they're a good general-purpose source for the kind of reference boundaries you'll often need to visualize or subset other spatial data against.

### Key functions:

`ne_countries()`: Returns country boundaries as an sf object, with a `continent` column you can filter or group by.

`ne_states()`: Returns state/province boundaries for a given country (from rnaturalearthhires).

Example:

```{r}
library(rnaturalearth)
library(rnaturalearthhires)

world <- ne_countries(scale = "medium", returnclass = "sf")

# Get New York state from the US
us_states <- ne_states(country = "United States of America", returnclass = "sf")
new_york <- us_states %>% filter(name == "New York")
new_york %>% ggplot() + geom_sf()
```

Get a simple feature of every country in South America, and one that unions them into a single South America polygon:

```{r}
south_america_countries <- world %>% filter(continent == "South America")
south_america <- south_america_countries %>% st_union() %>% st_as_sf()
south_america %>% ggplot() + geom_sf()
```

### How many features does south_america_countries have, versus south_america? Why?

```{r}
# Compare nrow() of the two objects here
# ...
```

Re-project south_america to match fluxnet_ch4_sf:

```{r}
aoi_south_america <- st_transform(south_america, crs = 4087)
aoi_south_america %>% ggplot() + geom_sf()
```

Create a simple feature of Brazil and re-project it to match fluxnet_ch4_sf:

```{r}
# Create your simple feature object here
# ...
```

The functions `st_intersects()`, `st_disjoint()`, `st_touches()`, `st_crosses()`, `st_within()`, `st_contains()`, `st_overlaps()`, `st_equals()`, `st_covers()`, `st_covered_by()`, `st_equals_exact()` and `st_is_within_distance()` all return a sparse matrix with matching (TRUE) indexes, or a full logical matrix:

### How many towers are in South America?

```{r}
st_intersects(aoi_south_america, fluxnet_ch4_sf)
st_intersects(aoi_south_america, fluxnet_ch4_sf, sparse = FALSE)
```

### How many towers are in Brazil?

```{r}
# Find out how many towers are in Brazil here
# ...
```

Where possible geometric operations such as `st_distance()`, `st_length()` and `st_area()` report results with a units attribute appropriate for the CRS.

Calculate the area of South America:

```{r}
aoi_south_america$area <- st_area(aoi_south_america)
aoi_south_america$area
```

Calculate the area of Brazil:

```{r}
# Calculate the area here
# ...
```

### Visualize the global distribution of towers:

First create a simple feature for every country in Europe, Asia, the Americas, Africa, and Oceania (i.e. all inhabited continents except Antarctica):

```{r}
aoi_terrestrial <- world %>% filter(continent %in% c("Europe", "Asia", "North America", "South America", "Africa", "Oceania"))

aoi_terrestrial %>% ggplot() + geom_sf()
```

Look at the CRS:

```{r}
aoi_terrestrial
```

Re-project the polygon to match fluxnet_ch4_sf:

```{r}
aoi_terrestrial <- aoi_terrestrial %>% st_transform(4087)

aoi_terrestrial
```

Visualize the shapefile you created:

```{r}
aoi_terrestrial %>% ggplot() + geom_sf()
```

Use ggplot to visualize the global distribution of Fluxnet CH4 sites:

```{r}
ggplot() + geom_sf(data = aoi_terrestrial) + geom_sf(data = fluxnet_ch4_sf)
```

Extract the country from the world simple feature into fluxnet_ch4_sf. A straight `st_intersection()` can return more or fewer rows than you started with (a point can sit exactly on a shared border and match two polygons, or fall outside all of them, e.g. small islands at this map resolution), so join back on `SITE_ID` instead of assigning the result directly:

```{r}
country_lookup <- st_intersection(aoi_terrestrial, fluxnet_ch4_sf) %>%
  st_drop_geometry() %>%
  select(SITE_ID, name) %>%
  distinct(SITE_ID, .keep_all = TRUE)

fluxnet_ch4_sf <- fluxnet_ch4_sf %>%
  left_join(country_lookup, by = "SITE_ID") %>%
  rename(country = name)
```

Explore the Fluxnet CH4 sites:

```{r}
names(fluxnet_ch4_sf)

ggplot(data = fluxnet_ch4_sf) + geom_point(aes(x = MAT, y = MAP))

ggplot(data = fluxnet_ch4_sf) + geom_point(aes(x = MAT, y = MAP, col = IGBP))

fluxnet_ch4_sf$IGBP <- as.factor(fluxnet_ch4_sf$IGBP)
summary(fluxnet_ch4_sf$IGBP)
```

### Writing files using st_write:

When writing, you can use the following arguments to control update and delete: update=TRUE causes an existing data source to be updated, if it exists; this option is by default TRUE for all database drivers, where the database is updated by adding a table.

delete_layer=TRUE causes st_write to try to open the data source and delete the layer; no errors are given if the data source is not present, or the layer does not exist in the data source.

delete_dsn=TRUE causes st_write to delete the data source when present, before writing the layer in a newly created data source. No error is given when the data source does not exist. This option should be handled with care, as it may wipe complete directories or databases.

```{r}
#| eval: false
write_sf(fluxnet_ch4_sf, "data/products/fluxnet_ch4.shp")
```

It is possible to create data.frame objects with geometry list-columns that are not of class sf by:

```{r}
fluxnet_ch4_df <- as.data.frame(fluxnet_ch4_sf)
```

Check the class:

```{r}
class(fluxnet_ch4_df)
```

Such objects: no longer register which column is the geometry list-column, no longer have a plot method, and lack all of the other dedicated methods listed above for class sf. To write this object:

```{r}
#| eval: false
write.csv(fluxnet_ch4_df, "data/fluxnet_ch4.csv")
```

### Additional Reading:

S. Scheider, B. Gräler, E. Pebesma, C. Stasch, 2016. Modelling spatio-temporal information generation. Int J of Geographic Information Science, 30 (10), 1980-2008. (open access) Stasch, C., S. Scheider, E. Pebesma, W. Kuhn, 2014. Meaningful Spatial Prediction and Aggregation. Environmental Modelling & Software, 51, (149–165, open access).

# Assignment:

Write a 3-page report on the distribution of tower sites discussing the strengths and weakness of the current tower representation. Please create 2 visualizations. You are welcome to use any additional data.
