---
title: "Vector Data: Mapping Gaps in the Global Methane Tower Network"
editor:
  mode: source
---

## Mapping methane observations

Models are only as good as the places where we collect data. FLUXNET towers provide direct measurements of ecosystem-atmosphere exchange, and FLUXNET-CH4 towers are especially important for understanding methane emissions from natural ecosystems. But the tower network is not evenly distributed across the planet.

In this workshop, we will use vector data to ask:

> If FLUXNET CH4 towers are used to understand natural methane emissions, where are we observing the planet, and where are we guessing?

By the end of the workshop, you will be able to:

1.  Convert tabular coordinates into simple feature objects.
2.  Inspect geometries and coordinate reference systems.
3.  Transform vector data into an appropriate projected CRS.
4.  Measure distances among point locations.
5.  Use country and continent polygons from `rnaturalearth`.
6.  Use spatial predicates and spatial joins to count towers inside regions.
7.  Calculate polygon area and tower density.
8.  Evaluate basic representativeness of the CH4 tower network.
9.  Write spatial outputs for later workshops.

### Install libraries for this workshop

```{r}
#| eval: false
install.packages("sf")
install.packages("rnaturalearth")
install.packages("tidyverse")
install.packages("ggplot2")
```

### Load the required libraries

```{r}
#| output: false
library(sf)
library(rnaturalearth)
library(tidyverse)
library(ggplot2)
```

# Data

Import `FluxNet_Sites_2024.csv`. This table was created from the FLUXNET site list found at <https://fluxnet.org/sites/site-list-and-pages/?view=table>.

```{r}
fluxnet_sites <- read.csv("data/FluxNet_Sites_2024.csv")
```

This dataset includes:

| Column Name | Description |
|---------------------------:|--------------------------------------------|
| SITE_ID | Unique site ID |
| SITE_NAME | Site name |
| FLUXNET2015 | License information for FLUXNET2015 |
| FLUXNET-CH4 | License information for FLUXNET-CH4 |
| LOCATION_LAT | Latitude |
| LOCATION_LONG | Longitude |
| LOCATION_ELEV | Elevation in meters |
| IGBP | Vegetation type |
| MAT | Mean annual temperature in degrees C |
| MAP | Mean annual precipitation in mm |

Take a look at the file:

```{r}
#| eval: false
View(fluxnet_sites)
```

Start with a non-spatial view of the tower locations. At this stage, latitude and longitude are just numbers:

```{r}
ggplot(data = fluxnet_sites,
       aes(x = LOCATION_LONG, y = LOCATION_LAT)) +
  geom_point() +
  labs(title = "FLUXNET tower locations as longitude and latitude",
       x = "Longitude",
       y = "Latitude")
```

Look at the climate and elevation conditions represented by the full tower list:

```{r}
ggplot(data = fluxnet_sites, aes(x = LOCATION_ELEV)) +
  geom_histogram(color = "black", fill = "white") +
  labs(title = "Elevation of FLUXNET tower sites",
       x = "Elevation (m)",
       y = "Count")

ggplot(data = fluxnet_sites, aes(x = MAT)) +
  geom_histogram(color = "black", fill = "white") +
  labs(title = "Mean annual temperature of FLUXNET tower sites",
       x = "Mean annual temperature (degrees C)",
       y = "Count")

ggplot(data = fluxnet_sites, aes(x = MAP)) +
  geom_histogram(color = "black", fill = "white") +
  labs(title = "Mean annual precipitation of FLUXNET tower sites",
       x = "Mean annual precipitation (mm)",
       y = "Count")
```

# From tables to simple features

We are interested in the sites with methane data. The column `FLUXNET-CH4` is read into R as `FLUXNET.CH4`, because R replaces the hyphen with a dot.

```{r}
fluxnet_ch4 <- fluxnet_sites %>%
  filter(FLUXNET.CH4 != "")
```

How many total FLUXNET sites are in the table? How many have CH4 data?

```{r}
nrow(fluxnet_sites)
nrow(fluxnet_ch4)
```

This object is still a data frame:

```{r}
class(fluxnet_ch4)
```

Latitude and longitude become spatial data when we tell R which columns contain coordinates and what coordinate reference system they use. A few rows in the full FLUXNET table may be missing coordinates, so remove those before converting the table to points. These coordinates are in WGS84 longitude/latitude, EPSG 4326.

```{r}
fluxnet_sites_located <- fluxnet_sites %>%
  filter(!is.na(LOCATION_LONG), !is.na(LOCATION_LAT))

fluxnet_ch4_located <- fluxnet_ch4 %>%
  filter(!is.na(LOCATION_LONG), !is.na(LOCATION_LAT))
```

```{r}
fluxnet_sites_sf <- st_as_sf(x = fluxnet_sites_located,
                              coords = c("LOCATION_LONG", "LOCATION_LAT"),
                              crs = 4326)

fluxnet_ch4_sf <- st_as_sf(x = fluxnet_ch4_located,
                            coords = c("LOCATION_LONG", "LOCATION_LAT"),
                            crs = 4326)
```

Now map the full FLUXNET network and the CH4 subset:

```{r}
ggplot() +
  geom_sf(data = fluxnet_sites_sf, color = "gray75", size = 1) +
  geom_sf(data = fluxnet_ch4_sf, color = "red", size = 1.6) +
  labs(title = "FLUXNET CH4 towers are a subset of the full FLUXNET network")
```

Check the class and geometry:

```{r}
class(fluxnet_ch4_sf)
fluxnet_ch4_sf$geometry
print(fluxnet_ch4_sf, n = 3)
```

Simple features describe real-world objects with both attributes and geometry. In this case, the attributes describe tower names, climate, elevation, and vegetation type, while the geometry records each tower's location.

# Geometry checks and coordinate reference systems

`st_is_valid()` checks whether geometries are valid. Points are usually simple, but it is still a good habit to check spatial data before using it.

```{r}
st_is_valid(fluxnet_ch4_sf)
```

Coordinate reference systems (CRS) define how coordinates map to locations on Earth. Longitude/latitude coordinates are useful for storing global locations, but projected coordinates are better when measuring area and distance. We will transform the tower data to World Equidistant Cylindrical, EPSG 4087, so distance and area operations use meters.

```{r}
fluxnet_sites_sf <- st_transform(fluxnet_sites_sf, crs = 4087)
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = 4087)

fluxnet_ch4_sf
```

# Are methane towers clustered?

If methane towers are clustered, models may describe those places well while leaving other climate or ecosystem conditions underrepresented. Use `st_distance()` to calculate the distance among CH4 towers, then identify each tower's nearest neighbor.

```{r}
distance_matrix_km <- as.matrix(st_distance(fluxnet_ch4_sf)) / 1000
diag(distance_matrix_km) <- NA

fluxnet_ch4_sf$nearest_neighbor_km <- apply(distance_matrix_km, 1, min, na.rm = TRUE)
```

Map nearest-neighbor distance. Larger values identify more isolated towers:

```{r}
ggplot(data = fluxnet_ch4_sf) +
  geom_sf(aes(color = nearest_neighbor_km), size = 2) +
  labs(title = "Nearest CH4 tower distance",
       color = "Nearest tower (km)")
```

Look at the distribution:

```{r}
ggplot(data = fluxnet_ch4_sf,
       aes(x = nearest_neighbor_km)) +
  geom_histogram(color = "black", fill = "white") +
  labs(title = "Distribution of nearest-neighbor distances",
       x = "Distance to nearest CH4 tower (km)",
       y = "Count")
```

Which towers are most isolated?

```{r}
fluxnet_ch4_sf %>%
  st_drop_geometry() %>%
  select(SITE_ID, SITE_NAME, IGBP, MAT, MAP, nearest_neighbor_km) %>%
  arrange(desc(nearest_neighbor_km)) %>%
  head()
```

# Add country and continent context

Point locations become more useful when we attach contextual information. We will get country polygons from `rnaturalearth` and spatially join each tower to a country and continent.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent %in% c("Europe", "Asia", "North America",
                          "South America", "Africa", "Oceania")) %>%
  st_transform(crs = 4087) %>%
  st_make_valid()
```

Map the global land polygons:

```{r}
ggplot(data = world) +
  geom_sf(fill = "gray95", color = "gray70") +
  labs(title = "Country polygons from rnaturalearth")
```

Attach country and continent to all FLUXNET sites and to CH4 sites:

```{r}
world_context <- world %>%
  select(country = name, continent, geometry)

fluxnet_sites_sf <- fluxnet_sites_sf %>%
  st_join(world_context, join = st_within)

fluxnet_ch4_sf <- fluxnet_ch4_sf %>%
  st_join(world_context, join = st_within)
```

Look at the new columns:

```{r}
fluxnet_ch4_sf %>%
  st_drop_geometry() %>%
  select(SITE_ID, SITE_NAME, country, continent) %>%
  head()
```

Map the CH4 tower network with country context in the background:

```{r}
ggplot() +
  geom_sf(data = world, fill = "gray95", color = "white") +
  geom_sf(data = fluxnet_ch4_sf, aes(color = continent), size = 1.8) +
  labs(title = "Global distribution of FLUXNET CH4 towers",
       color = "Continent")
```

# Case study: South America and Brazil

Tropical and subtropical wetlands are important to the global methane cycle. South America is a useful case study because it includes large wetland systems and major climate gradients.

Create a simple feature object for South America, and a second object that unions all South American countries into one polygon:

```{r}
south_america_countries <- world %>%
  filter(continent == "South America")

south_america <- south_america_countries %>%
  st_union() %>%
  st_as_sf() %>%
  rename(geometry = x)

ggplot(data = south_america) +
  geom_sf() +
  labs(title = "South America as one polygon")
```

How many features does `south_america_countries` have, versus `south_america`? Why?

```{r}
nrow(south_america_countries)
nrow(south_america)
```

Create a Brazil polygon:

```{r}
brazil <- world %>%
  filter(name == "Brazil")

ggplot(data = brazil) +
  geom_sf() +
  labs(title = "Brazil")
```

The functions `st_intersects()`, `st_within()`, `st_contains()`, `st_disjoint()`, and related predicates tell us whether spatial features have a specific relationship. Use `st_intersects()` to identify CH4 towers in South America:

```{r}
south_america_intersections <- st_intersects(south_america, fluxnet_ch4_sf)
south_america_intersections

south_america_tower_count <- lengths(south_america_intersections)
south_america_tower_count
```

Count towers in Brazil:

```{r}
brazil_intersections <- st_intersects(brazil, fluxnet_ch4_sf)
brazil_tower_count <- lengths(brazil_intersections)
brazil_tower_count
```

Calculate area and tower density. Area is easier to interpret in million square kilometers:

```{r}
south_america_area_million_km2 <- as.numeric(st_area(south_america)) / 1e12
brazil_area_million_km2 <- as.numeric(st_area(brazil)) / 1e12

south_america_density <- south_america_tower_count / south_america_area_million_km2
brazil_density <- brazil_tower_count / brazil_area_million_km2

south_america_density
brazil_density
```

Create a small comparison table:

```{r}
regional_density <- tibble(
  region = c("South America", "Brazil"),
  ch4_towers = c(south_america_tower_count, brazil_tower_count),
  area_million_km2 = c(south_america_area_million_km2, brazil_area_million_km2),
  towers_per_million_km2 = c(south_america_density, brazil_density)
)

regional_density
```

# Basic representativeness analysis

Now we can ask whether FLUXNET CH4 sites are representative of the larger FLUXNET network. This is not a complete representativeness analysis, but it is a useful first pass. We will compare:

-   geographic representation by continent;
-   ecosystem representation by `IGBP`;
-   climate space coverage using mean annual temperature (`MAT`) and precipitation (`MAP`);
-   elevation coverage.

### Representation by continent

Count all FLUXNET sites by continent and CH4 sites by continent:

```{r}
all_sites_by_continent <- fluxnet_sites_sf %>%
  st_drop_geometry() %>%
  count(continent, name = "all_fluxnet_sites")

ch4_sites_by_continent <- fluxnet_ch4_sf %>%
  st_drop_geometry() %>%
  count(continent, name = "ch4_sites")
```

Join the summaries and calculate the fraction of FLUXNET sites that have CH4 data:

```{r}
representation_by_continent <- all_sites_by_continent %>%
  full_join(ch4_sites_by_continent, by = "continent") %>%
  mutate(ch4_sites = replace_na(ch4_sites, 0),
         ch4_fraction = ch4_sites / all_fluxnet_sites) %>%
  arrange(desc(ch4_fraction))

representation_by_continent
```

Visualize the result:

```{r}
ggplot(data = representation_by_continent,
       aes(x = reorder(continent, ch4_fraction), y = ch4_fraction)) +
  geom_col() +
  coord_flip() +
  labs(title = "Fraction of FLUXNET sites with CH4 data by continent",
       x = "Continent",
       y = "CH4 sites / all FLUXNET sites")
```

### Representation by ecosystem type

Compare ecosystem representation using `IGBP`:

```{r}
all_sites_by_igbp <- fluxnet_sites_sf %>%
  st_drop_geometry() %>%
  count(IGBP, name = "all_fluxnet_sites")

ch4_sites_by_igbp <- fluxnet_ch4_sf %>%
  st_drop_geometry() %>%
  count(IGBP, name = "ch4_sites")

representation_by_igbp <- all_sites_by_igbp %>%
  full_join(ch4_sites_by_igbp, by = "IGBP") %>%
  mutate(ch4_sites = replace_na(ch4_sites, 0),
         ch4_fraction = ch4_sites / all_fluxnet_sites) %>%
  arrange(desc(ch4_sites))

representation_by_igbp
```

Visualize CH4 tower counts by ecosystem type:

```{r}
ggplot(data = representation_by_igbp,
       aes(x = reorder(IGBP, ch4_sites), y = ch4_sites)) +
  geom_col() +
  coord_flip() +
  labs(title = "FLUXNET CH4 tower count by ecosystem type",
       x = "IGBP",
       y = "CH4 tower count")
```

### Climate space coverage

A spatial network can be geographically widespread but still miss important climate conditions. Compare all FLUXNET sites to CH4 sites in climate space:

```{r}
ggplot() +
  geom_point(data = fluxnet_sites,
             aes(x = MAT, y = MAP),
             color = "gray70") +
  geom_point(data = fluxnet_ch4,
             aes(x = MAT, y = MAP, color = IGBP)) +
  labs(title = "Climate space represented by FLUXNET CH4 towers",
       subtitle = "Gray points show all FLUXNET sites; colored points show CH4 sites",
       x = "Mean annual temperature (degrees C)",
       y = "Mean annual precipitation (mm)",
       color = "IGBP")
```

Compare elevation distributions:

```{r}
fluxnet_site_type <- bind_rows(
  fluxnet_sites %>%
    mutate(site_type = "All FLUXNET sites"),
  fluxnet_ch4 %>%
    mutate(site_type = "CH4 sites")
)

ggplot(data = fluxnet_site_type,
       aes(x = LOCATION_ELEV, fill = site_type)) +
  geom_density(alpha = 0.45) +
  labs(title = "Elevation conditions represented by FLUXNET and CH4 towers",
       x = "Elevation (m)",
       y = "Density",
       fill = "Site type")
```

### Interpreting representation

Use the tables and figures above to answer:

1.  Which continents have many CH4 towers?
2.  Which continents have a low fraction of CH4 towers relative to all FLUXNET sites?
3.  Which ecosystem types dominate the CH4 network?
4.  Are CH4 towers missing any obvious climate space in the `MAT` and `MAP` plot?
5.  If you could add one new tower, where would you prioritize it?

# Writing spatial files

We will use the CH4 tower simple feature in later workshops. Save a shapefile version:

```{r}
#| eval: false
write_sf(fluxnet_ch4_sf, "data/products/fluxnet_ch4.shp")
```

It is also possible to convert a simple feature object back to a data frame:

```{r}
fluxnet_ch4_df <- as.data.frame(fluxnet_ch4_sf)
```

Check the class:

```{r}
class(fluxnet_ch4_df)
```

Once the object is a data frame, it no longer has dedicated `sf` methods. The geometry column is still present, but R no longer treats the object as a spatial layer. Write a CSV version:

```{r}
#| eval: false
write.csv(fluxnet_ch4_df, "data/fluxnet_ch4.csv")
```

### Additional reading

S. Scheider, B. Graler, E. Pebesma, C. Stasch, 2016. Modelling spatio-temporal information generation. International Journal of Geographic Information Science, 30 (10), 1980-2008. Stasch, C., S. Scheider, E. Pebesma, W. Kuhn, 2014. Meaningful Spatial Prediction and Aggregation. Environmental Modelling & Software, 51, 149-165.

::: {.callout-tip}
## Course product
This workshop produces a methane tower representation audit. Use the standard deliverable shape from the Final Project Roadmap: one clear question, one reproducibility note, at least one table, one or two figures, one interpretation paragraph, and one limitation or next-step paragraph.
:::

# Assignment

Write a short FLUXNET CH4 representation audit. Your report should answer:

1.  Where are CH4 towers geographically concentrated?
2.  Which continents, countries, ecosystem types, or climate spaces seem underrepresented?
3.  What do the nearest-neighbor distances suggest about clustering or isolation in the network?
4.  Based on your analysis, where would you place one new methane tower and why?

Include at least two maps and one non-map figure. You are welcome to use additional data, but your report should include evidence from the spatial and representativeness analyses in this workshop.
