---
title: "Spatial Model Visualization: Scaling Methane Flux in Space"
editor:
  mode: source
---

## From flux density to emissions

In the Random Forest and Sensitivity Analysis workshops, we built and stress-tested a methane flux model. In this final workshop, we will project a model across Connecticut and turn gridded flux predictions into a methane emissions budget.

The key idea is a unit conversion:

> methane flux density x inundated area = methane emissions

Our random forest predicts methane flux as `g C m-2 month-1`. That is a rate per square meter. A map of predicted flux is not yet an emissions budget. To estimate emissions, we need to multiply by the area where methane-producing wetlands are present.

That is where the **Wetland Area and Dynamics for Methane Modeling (WAD2M)** product enters. WAD2M provides monthly wetland/inundation fractions on a global 0.25 degree grid. Zhang et al. (2021) describe WAD2M version 1.0 as a monthly product for 2000-2018 built for large-scale methane modeling. The NASA UpCH4 methane emissions product uses WAD2M to weight gridded methane flux predictions by fractional wetland extent before converting them to by-grid-cell emissions.

Useful references:

-   WAD2M paper: [Zhang et al. 2021, Earth System Science Data](https://doi.org/10.5194/essd-13-2001-2021)
-   WAD2M metadata summary: [EPA HERO record](https://hero.epa.gov/reference/8047289/)
-   Example methane upscaling product using WAD2M: [NASA Open Data Portal, UpCH4](https://data.nasa.gov/dataset/global-wetland-methane-emissions-derived-from-fluxnet-and-the-upch4-model-2001-2018-c7437)

::: {.callout-warning}
## Data dependency
This workshop reuses the ERA5 climate rasters saved in `data_integration.qmd`: `data/products/era5_ppt_monthly.tif` and `data/products/era5_tmean_monthly.tif`.

For WAD2M, this workshop expects a Connecticut-ready 2018 monthly inundation fraction raster at `data/products/wad2m_inundation_fraction_ct_2018.tif`. If that file is not present, the code creates a clearly labeled seasonal classroom proxy from the provided MODIS land-cover raster so the unit-conversion workflow still runs.
:::

By the end of this workshop, you will be able to:

1.  Identify the variables, units, names, and classes needed for spatial prediction.
2.  Align predictor rasters so they have the same CRS, resolution, extent, and layer names.
3.  Predict monthly methane flux density in space.
4.  Explain what WAD2M inundation fraction represents.
5.  Convert flux density to cell-level monthly and annual emissions.
6.  Compare budgets with and without inundation-fraction scaling.
7.  Extract predictions and emissions for the Yale-Myers Forest grid cell.
8.  Test how sensitive the budget is to inundation-area uncertainty.
9.  Report methane emissions in `g C`, `Tg C`, and `Tg CH4`.

### Load libraries

```{r}
#| output: false
library(randomForest)
library(terra)
library(tidyverse)
```

# 1. List model variables and units

Load the model. This workshop uses the original spatial-projection model because it has three predictors that are available as rasters for Connecticut: precipitation, temperature, and an upland/inundated land-cover indicator.

```{r}
load(file = "data/final_model.RDATA")

# The saved model object is called FCH4_F_gC.rf; alias it to match this course's naming convention.
fch4_rf <- FCH4_F_gC.rf
```

Inspect the model formula and variable classes:

```{r}
formula(fch4_rf)
fch4_rf

class(train$P_F)
class(train$TA_F)
class(train$Upland)
```

To project this model in space, the raster stack must contain:

| Model variable | Meaning | Unit or class | Spatial source |
|---|---|---|---|
| `P_F` | Monthly total precipitation | mm month-1 | ERA5 |
| `TA_F` | Monthly mean air temperature | degrees C | ERA5 |
| `Upland` | Ecosystem indicator | factor with levels `inundated`, `upland` | MODIS land cover |

The model prediction has units of `g C m-2 month-1`.

::: {.callout-note}
## Which random forest should you project?
This workshop uses the simple spatial model because its predictors are easy to map for every grid cell in Connecticut. The updated Random Forest workshop may save a stronger model at `data/products/random_forest_model.RDATA`, but that model can only be projected if every selected predictor has a spatial raster with the same name and class expected by the model.

For example, a model using `NDVI`, `EVI`, `VPD_F`, or `IGBP_model` requires matching spatial layers for those variables. A good final project model is not only the model with the best test metrics; it is also a model whose predictors can be mapped honestly.
:::

Check whether the updated random forest exists and inspect its predictors:

```{r}
if (file.exists("data/products/random_forest_model.RDATA")) {
  load("data/products/random_forest_model.RDATA")
  updated_model_predictors <- final_predictors
} else {
  updated_model_predictors <- NA_character_
}

updated_model_predictors
```

Reload the simple spatial model for the rest of this workshop:

```{r}
load(file = "data/final_model.RDATA")
fch4_rf <- FCH4_F_gC.rf
```

# 2. Build the Connecticut prediction grid

Use the provided Connecticut MODIS raster as the Connecticut template. This gives us an area of interest and a target grid without requiring students to download a state boundary file.

```{r}
igbp_ct <- terra::rast("data/MODIS_IGBP_2001-2022_CT.tif")
ct_template <- igbp_ct[[2018 - 2001 + 1]]
ct_extent <- ext(ct_template)

terra::plot(ct_template, main = "Connecticut MODIS land-cover template")
```

Load the ERA5 rasters saved in Data Integration, keep only the 2018 layers, then crop and mask them to Connecticut:

```{r}
era5_ppt_all <- terra::rast("data/products/era5_ppt_monthly.tif")
era5_tmean_all <- terra::rast("data/products/era5_tmean_monthly.tif")

ct_era5_ppt <- era5_ppt_all[[grepl("2018", names(era5_ppt_all))]] %>%
  crop(ct_extent)

ct_era5_tmean <- era5_tmean_all[[grepl("2018", names(era5_tmean_all))]] %>%
  crop(ct_extent)

ct_era5_tmean
```

Save the cropped climate layers if you want to reuse them:

```{r}
#| eval: false
writeRaster(ct_era5_tmean, "data/products/era5_tmean_ct_2018.tif", overwrite = TRUE)
writeRaster(ct_era5_ppt, "data/products/era5_ppt_ct_2018.tif", overwrite = TRUE)
```

# 3. Build the land-cover predictor

Inspect the MODIS IGBP layer for Connecticut:

```{r}
igbp_ct

terra::plot(igbp_ct[[1]])
```

The MODIS IGBP classes include forest, shrubland, savanna, grassland, wetlands, cropland, urban land, snow/ice, barren land, and water. For this model, we simplify those classes into `Upland`.

```{r}
igbp_ct_r <- igbp_ct

igbp_ct_r[igbp_ct_r == 0] <- NA
igbp_ct_r[igbp_ct_r == 1] <- 1
igbp_ct_r[igbp_ct_r == 2] <- 1
igbp_ct_r[igbp_ct_r == 3] <- 1
igbp_ct_r[igbp_ct_r == 4] <- 1
igbp_ct_r[igbp_ct_r == 5] <- 1
igbp_ct_r[igbp_ct_r == 6] <- 1
igbp_ct_r[igbp_ct_r == 7] <- 1
igbp_ct_r[igbp_ct_r == 8] <- 1
igbp_ct_r[igbp_ct_r == 9] <- 1
igbp_ct_r[igbp_ct_r == 10] <- 1
igbp_ct_r[igbp_ct_r == 11] <- 0
igbp_ct_r[igbp_ct_r == 12] <- NA
igbp_ct_r[igbp_ct_r == 13] <- NA
igbp_ct_r[igbp_ct_r == 14] <- NA
igbp_ct_r[igbp_ct_r == 15] <- 0
igbp_ct_r[igbp_ct_r == 16] <- 1
igbp_ct_r[igbp_ct_r == 17] <- 0
```

Format the `Upland` raster as a factor:

```{r}
factors_df <- data.frame(id = c(0, 1), cover = c("inundated", "upland"))

for (i in seq_len(nlyr(igbp_ct_r))) {
  levels(igbp_ct_r[[i]]) <- factors_df
}
```

Subset the 2018 layer:

```{r}
igbp_ct_upland <- igbp_ct_r[[2018 - 2001 + 1]]
names(igbp_ct_upland) <- "Upland"

terra::plot(igbp_ct_upland)
```

Use the land-cover grid as the target grid, then resample the climate layers to match it:

```{r}
igbp_ct_upland <- terra::project(igbp_ct_upland, ct_era5_ppt)

ct_era5_tmean_resample <- resample(ct_era5_tmean, igbp_ct_upland)
ct_era5_ppt_resample <- resample(ct_era5_ppt, igbp_ct_upland)
```

Check that the layers now align:

```{r}
compareGeom(igbp_ct_upland, ct_era5_tmean_resample, ct_era5_ppt_resample)
```

# 4. Introduce WAD2M inundation fraction

WAD2M is not a model predictor in this simple random forest. It is used after prediction to scale fluxes by wetland or inundated area.

The values should be fractions between 0 and 1:

-   `0` means no WAD2M wetland/inundated area in the grid cell.
-   `0.25` means 25% of the grid cell is wetland/inundated area.
-   `1` means the whole grid cell is wetland/inundated area.

## Prepare WAD2M for Connecticut

The exact WAD2M download workflow depends on the format you obtain from the data portal. The key preparation steps are always the same:

1.  Load the global WAD2M inundation-fraction product.
2.  Select the 12 monthly layers for 2018.
3.  Crop the product to Connecticut.
4.  Project and resample it to the same grid as the model predictions.
5.  Save the prepared raster as `data/products/wad2m_inundation_fraction_ct_2018.tif`.

The following code is a template for preparing a local WAD2M NetCDF or GeoTIFF file after download:

```{r}
#| eval: false
wad2m_global <- rast("path/to/WAD2M_monthly_inundation_fraction.nc")

# Replace this pattern with the actual layer names in your WAD2M file.
wad2m_2018 <- wad2m_global[[grepl("2018", names(wad2m_global))]]

wad2m_ct_2018 <- wad2m_2018 %>%
  crop(ct_extent) %>%
  project(igbp_ct_upland) %>%
  resample(igbp_ct_upland, method = "bilinear") %>%
  clamp(lower = 0, upper = 1, values = TRUE)

names(wad2m_ct_2018) <- sprintf("wad2m_fraction_m%02d", 1:12)

writeRaster(wad2m_ct_2018,
            "data/products/wad2m_inundation_fraction_ct_2018.tif",
            overwrite = TRUE)
```

## If your target year is not available

WAD2M may not cover the year you want to model. For example, WAD2M version 1.0 covers 2000-2018, but your final project might need a later year. In that case, do not leave inundation fraction out of the budget. Instead, calculate a **mean monthly inundation climatology** from the WAD2M years that are available.

The idea is:

-   average all available January layers to create a typical January inundation fraction;
-   average all available February layers to create a typical February inundation fraction;
-   repeat for all 12 months;
-   use those 12 monthly averages to scale the target year's flux predictions.

This preserves seasonal wetland-area dynamics, even though it cannot represent the exact wetland extent of the missing year.

```{r}
#| eval: false
target_year <- 2021

wad2m_global <- rast("path/to/WAD2M_monthly_inundation_fraction.nc")

wad2m_dates <- tibble(
  layer = seq_len(nlyr(wad2m_global)),
  layer_name = names(wad2m_global)
) %>%
  mutate(
    year = as.integer(str_extract(layer_name, "\\d{4}")),
    month = as.integer(str_extract(layer_name, "(?<=-)\\d{2}|(?<=_)\\d{2}$"))
  )

target_layers <- wad2m_dates %>%
  filter(year == target_year)

if (nrow(target_layers) == 12) {
  wad2m_target <- wad2m_global[[target_layers$layer]]
} else {
  wad2m_monthly_climatology <- map(1:12, function(month_id) {
    month_layers <- wad2m_dates %>%
      filter(month == month_id) %>%
      pull(layer)

    mean(wad2m_global[[month_layers]], na.rm = TRUE)
  })

  wad2m_target <- rast(wad2m_monthly_climatology)
}

wad2m_target_ct <- wad2m_target %>%
  crop(ct_extent) %>%
  project(igbp_ct_upland) %>%
  resample(igbp_ct_upland, method = "bilinear") %>%
  clamp(lower = 0, upper = 1, values = TRUE)

names(wad2m_target_ct) <- sprintf("wad2m_fraction_m%02d", 1:12)

writeRaster(wad2m_target_ct,
            sprintf("data/products/wad2m_inundation_fraction_ct_%s.tif", target_year),
            overwrite = TRUE)
```

::: {.callout-note}
## Report this clearly
If you use a monthly climatology because the target year is unavailable, say so in your methods. Write something like: "Because WAD2M was unavailable for the target year, monthly inundation fractions were estimated using the mean WAD2M fraction for each calendar month across available years."
:::

Load the Connecticut 2018 WAD2M layer if it exists. The file should have 12 layers, one for each month of 2018. If it has one layer, the workflow will reuse that layer for all months. The fallback proxy below is only for teaching the unit conversion; replace it with real WAD2M before using the results scientifically.

```{r}
wad2m_path <- "data/products/wad2m_inundation_fraction_ct_2018.tif"

if (file.exists(wad2m_path)) {
  wad2m_fraction <- rast(wad2m_path)
  wad2m_source <- "WAD2M inundation fraction"
} else {
  natural_mask <- igbp_ct[[2018 - 2001 + 1]]
  natural_mask[natural_mask %in% c(0, 12, 13, 14)] <- NA
  natural_mask[!is.na(natural_mask)] <- 1

  seasonal_fraction <- c(0.04, 0.05, 0.07, 0.10, 0.12, 0.11,
                         0.09, 0.08, 0.07, 0.06, 0.05, 0.04)

  wad2m_fraction <- rast(lapply(seasonal_fraction, function(x) natural_mask * x))
  wad2m_source <- "seasonal classroom proxy, not WAD2M"
}

wad2m_source
```

Align the inundation fraction layer to the prediction grid:

```{r}
wad2m_fraction <- terra::project(wad2m_fraction, igbp_ct_upland)
wad2m_fraction <- terra::resample(wad2m_fraction, igbp_ct_upland, method = "bilinear")
wad2m_fraction <- terra::clamp(wad2m_fraction, lower = 0, upper = 1, values = TRUE)

if (nlyr(wad2m_fraction) == 1) {
  wad2m_fraction <- rep(wad2m_fraction, 12)
}

wad2m_fraction <- wad2m_fraction[[1:12]]
names(wad2m_fraction) <- sprintf("wad2m_fraction_m%02d", 1:12)

terra::plot(wad2m_fraction[[1]], main = "January inundation fraction")
```

Check the fraction values:

```{r}
global(wad2m_fraction, range, na.rm = TRUE)
global(wad2m_fraction, mean, na.rm = TRUE)
```

# 5. Make monthly flux predictions

Build a raster stack for January:

```{r}
model_rasters_m1 <- c(igbp_ct_upland,
                      ct_era5_tmean_resample[[1]],
                      ct_era5_ppt_resample[[1]])

names(model_rasters_m1) <- c("Upland", "TA_F", "P_F")
model_rasters_m1
```

Check that the raster classes match the training data:

```{r}
class(train$Upland)
levels(train$Upland)
model_rasters_m1$Upland
```

Predict January methane flux density:

```{r}
model_rasters_m1_pred <- terra::predict(object = model_rasters_m1, model = fch4_rf)
names(model_rasters_m1_pred) <- "F_CH4"
units(model_rasters_m1_pred) <- "g C m-2 month-1"

plot(model_rasters_m1_pred)
```

Predict all 12 months:

```{r}
monthly_predictions <- list()

for (i in seq_len(12)) {
  model_rasters <- c(igbp_ct_upland,
                     ct_era5_tmean_resample[[i]],
                     ct_era5_ppt_resample[[i]])

  names(model_rasters) <- c("Upland", "TA_F", "P_F")

  pred <- terra::predict(object = model_rasters, model = fch4_rf)
  names(pred) <- sprintf("F_CH4_m%02d", i)
  units(pred) <- "g C m-2 month-1"

  monthly_predictions[[i]] <- pred
}

predictions <- rast(monthly_predictions)
predictions
```

Save the monthly prediction rasters if you want to use them later:

```{r}
#| eval: false
predictions_dir <- file.path("data", "products", "predictions")
dir.create(predictions_dir, showWarnings = FALSE, recursive = TRUE)

for (i in seq_len(nlyr(predictions))) {
  writeRaster(predictions[[i]],
              file.path(predictions_dir, sprintf("model_pred_m%02d.tif", i)),
              overwrite = TRUE)
}
```

# 6. Convert flux density to emissions

The model predictions are flux densities:

`g C m-2 month-1`

To convert each grid cell to monthly emissions, multiply by inundated area:

`g C m-2 month-1 x m2 = g C month-1`

Keep the units visible as you work:

```{r}
unit_conversion_table <- tibble(
  quantity = c("predicted flux density",
               "cell area",
               "inundation fraction",
               "inundated area",
               "monthly emissions",
               "annual emissions",
               "annual emissions"),
  calculation = c("random forest prediction",
                  "terra::cellSize()",
                  "WAD2M value",
                  "cell area x inundation fraction",
                  "flux density x inundated area",
                  "sum monthly g C / 1e12",
                  "Tg C x 16 / 12"),
  unit = c("g C m-2 month-1",
           "m2 cell-1",
           "unitless",
           "m2 cell-1",
           "g C cell-1 month-1",
           "Tg C year-1",
           "Tg CH4 year-1")
)

unit_conversion_table
```

First calculate total cell area in square meters:

```{r}
cell_area_m2 <- cellSize(predictions[[1]], unit = "m")
names(cell_area_m2) <- "cell_area_m2"

plot(cell_area_m2)
```

Now calculate monthly inundated area:

```{r}
inundated_area_m2 <- wad2m_fraction * cell_area_m2
names(inundated_area_m2) <- sprintf("inundated_area_m2_m%02d", 1:12)

global(inundated_area_m2, sum, na.rm = TRUE)
```

Convert monthly flux density to monthly emissions:

```{r}
monthly_emissions_g_c <- predictions * inundated_area_m2
names(monthly_emissions_g_c) <- sprintf("emissions_g_c_m%02d", 1:12)
units(monthly_emissions_g_c) <- "g C month-1"

monthly_emissions_g_c
```

Compute annual emissions per grid cell:

```{r}
annual_emissions_g_c <- sum(monthly_emissions_g_c)
names(annual_emissions_g_c) <- "annual_emissions_g_c"
units(annual_emissions_g_c) <- "g C year-1"

plot(annual_emissions_g_c)
```

Convert `g C` to larger reporting units:

```{r}
annual_emissions_tg_c <- annual_emissions_g_c / 1e12
names(annual_emissions_tg_c) <- "annual_emissions_tg_c"
units(annual_emissions_tg_c) <- "Tg C year-1"

annual_emissions_tg_ch4 <- annual_emissions_tg_c * (16 / 12)
names(annual_emissions_tg_ch4) <- "annual_emissions_tg_ch4"
units(annual_emissions_tg_ch4) <- "Tg CH4 year-1"
```

The molecular-weight conversion is:

`Tg CH4 = Tg C x 16 / 12`

because methane is `CH4`, with molecular mass 16, and carbon has atomic mass 12.

# 7. Build the Connecticut methane budget

Sum across grid cells:

```{r}
ct_budget <- tibble(
  quantity = c("annual emissions", "annual emissions"),
  unit = c("Tg C year-1", "Tg CH4 year-1"),
  value = c(
    global(annual_emissions_tg_c, "sum", na.rm = TRUE)[1, 1],
    global(annual_emissions_tg_ch4, "sum", na.rm = TRUE)[1, 1]
  )
)

ct_budget
```

Create a monthly budget:

```{r}
monthly_budget <- tibble(
  month = 1:12,
  emissions_g_c = as.numeric(global(monthly_emissions_g_c, "sum", na.rm = TRUE)[1, ]),
  emissions_tg_c = emissions_g_c / 1e12,
  emissions_tg_ch4 = emissions_tg_c * (16 / 12),
  inundated_area_km2 = as.numeric(global(inundated_area_m2, "sum", na.rm = TRUE)[1, ]) / 1e6
)

monthly_budget
```

Plot the seasonal budget:

```{r}
monthly_budget %>%
  ggplot(aes(x = month, y = emissions_tg_ch4)) +
  geom_line() +
  geom_point() +
  labs(title = "Monthly Connecticut wetland methane emissions",
       x = "Month",
       y = "Tg CH4 month-1")
```

# 8. Extract Yale-Myers Forest

Now connect the spatial budget back to Yale-Myers Forest. These values represent the coarse model grid cell containing Yale-Myers, not a field-scale measurement of the forest.

```{r}
yale_myers <- vect(
  data.frame(
    site = "Yale-Myers Forest",
    lon = -72.123859,
    lat = 41.952909
  ),
  geom = c("lon", "lat"),
  crs = "EPSG:4326"
)

yale_myers <- project(yale_myers, predictions)

plot(predictions[[7]], main = "July predicted methane flux")
points(yale_myers, pch = 16, col = "red")
```

Extract monthly flux density, inundation fraction, inundated area, and emissions for the grid cell containing Yale-Myers:

```{r}
yale_flux <- terra::extract(predictions, yale_myers) %>%
  select(-ID) %>%
  pivot_longer(everything(),
               names_to = "flux_layer",
               values_to = "flux_g_c_m2_month")

yale_fraction <- terra::extract(wad2m_fraction, yale_myers) %>%
  select(-ID) %>%
  pivot_longer(everything(),
               names_to = "fraction_layer",
               values_to = "inundation_fraction")

yale_area <- terra::extract(inundated_area_m2, yale_myers) %>%
  select(-ID) %>%
  pivot_longer(everything(),
               names_to = "area_layer",
               values_to = "inundated_area_m2")

yale_emissions <- terra::extract(monthly_emissions_g_c, yale_myers) %>%
  select(-ID) %>%
  pivot_longer(everything(),
               names_to = "emissions_layer",
               values_to = "emissions_g_c_month")

yale_myers_monthly <- tibble(month = 1:12) %>%
  bind_cols(yale_flux, yale_fraction, yale_area, yale_emissions) %>%
  mutate(
    emissions_tg_c_month = emissions_g_c_month / 1e12,
    emissions_tg_ch4_month = emissions_tg_c_month * (16 / 12)
  ) %>%
  select(month,
         flux_g_c_m2_month,
         inundation_fraction,
         inundated_area_m2,
         emissions_g_c_month,
         emissions_tg_ch4_month)

yale_myers_monthly
```

Summarize the Yale-Myers grid cell:

```{r}
yale_myers_summary <- yale_myers_monthly %>%
  summarize(
    mean_flux_g_c_m2_month = mean(flux_g_c_m2_month, na.rm = TRUE),
    mean_inundation_fraction = mean(inundation_fraction, na.rm = TRUE),
    max_inundated_area_m2 = max(inundated_area_m2, na.rm = TRUE),
    annual_emissions_tg_ch4 = sum(emissions_tg_ch4_month, na.rm = TRUE)
  )

yale_myers_summary
```

Plot the seasonal signal for the Yale-Myers grid cell:

```{r}
yale_myers_monthly %>%
  ggplot(aes(x = month, y = emissions_tg_ch4_month)) +
  geom_line() +
  geom_point() +
  labs(title = "Yale-Myers grid-cell methane emissions",
       x = "Month",
       y = "Tg CH4 month-1")
```

# 9. Why WAD2M scaling matters

Compare the WAD2M-scaled budget to a naive budget that assumes every valid cell is 100% emitting wetland area:

```{r}
naive_monthly_emissions_g_c <- predictions * cell_area_m2
naive_annual_tg_ch4 <- sum(naive_monthly_emissions_g_c) / 1e12 * (16 / 12)
names(naive_annual_tg_ch4) <- "naive_annual_tg_ch4"

comparison_budget <- tibble(
  method = c("100% of each valid cell", "WAD2M-scaled inundated fraction"),
  annual_tg_ch4 = c(
    global(naive_annual_tg_ch4, "sum", na.rm = TRUE)[1, 1],
    global(annual_emissions_tg_ch4, "sum", na.rm = TRUE)[1, 1]
  )
) %>%
  mutate(percent_of_naive = 100 * annual_tg_ch4 / first(annual_tg_ch4))

comparison_budget
```

Plot the comparison:

```{r}
comparison_budget %>%
  ggplot(aes(x = method, y = annual_tg_ch4)) +
  geom_col() +
  coord_flip() +
  labs(title = "Effect of inundation-fraction scaling",
       x = NULL,
       y = "Annual emissions (Tg CH4 year-1)")
```

# 10. Inundation-area uncertainty

Wetland area is often one of the largest sources of uncertainty in methane budgets. Here we ask how much the annual Connecticut budget changes if inundation fraction is 50% lower or 50% higher than the mapped value.

```{r}
inundation_uncertainty <- tibble(
  scenario = c("low inundation", "mapped inundation", "high inundation"),
  multiplier = c(0.5, 1, 1.5)
)

budget_uncertainty <- inundation_uncertainty %>%
  mutate(annual_tg_ch4 = map_dbl(multiplier, function(multiplier_value) {
    adjusted_fraction <- clamp(wad2m_fraction * multiplier_value,
                               lower = 0,
                               upper = 1,
                               values = TRUE)
    adjusted_area <- adjusted_fraction * cell_area_m2
    adjusted_emissions <- predictions * adjusted_area
    adjusted_annual_tg_ch4 <- sum(adjusted_emissions) / 1e12 * (16 / 12)

    global(adjusted_annual_tg_ch4, "sum", na.rm = TRUE)[1, 1]
  }))

budget_uncertainty
```

Plot the uncertainty range:

```{r}
budget_uncertainty %>%
  ggplot(aes(x = scenario, y = annual_tg_ch4)) +
  geom_col() +
  labs(title = "Sensitivity of the methane budget to inundation area",
       x = NULL,
       y = "Annual emissions (Tg CH4 year-1)")
```

This is not a full uncertainty analysis. It isolates one uncertainty source: the area used to scale flux density into emissions. A final project could also vary the methane model, predictor rasters, or the wetland mask.

# 11. Save products

Save the annual emission maps if you want to use them in a final report:

```{r}
#| eval: false
writeRaster(annual_emissions_tg_c,
            "data/products/ct_annual_methane_emissions_tg_c_2018.tif",
            overwrite = TRUE)

writeRaster(annual_emissions_tg_ch4,
            "data/products/ct_annual_methane_emissions_tg_ch4_2018.tif",
            overwrite = TRUE)
```

# Assignment

For your final project, prepare a spatial methane visualization and budget. Include:

1.  Your model formula and the exact raster layer names needed for prediction.
2.  A table listing each spatial predictor, units, CRS, resolution, and source.
3.  A map of predicted methane flux density in `g C m-2 month-1`.
4.  A map or summary of the WAD2M inundation fraction used to scale the model.
5.  A unit-conversion table showing:
    -   flux density: `g C m-2 month-1`;
    -   cell area: `m2`;
    -   inundated area: `m2 x fraction`;
    -   monthly emissions: `g C month-1`;
    -   annual emissions: `Tg C year-1`;
    -   annual emissions: `Tg CH4 year-1`.
6.  A comparison between a naive 100% area budget and a WAD2M-scaled budget.
7.  A Yale-Myers or target-site extraction table.
8.  A simple inundation-area uncertainty analysis.
9.  One paragraph explaining why inundation fraction changes the interpretation of a spatial methane flux map.
