---
title: "Introduction to rasters with the terra package"
editor:
  mode: source
---

### Install libraries for this workshop:

```{r}
#| eval: false
install.packages("sf")
install.packages("terra")
install.packages("tidyverse")
install.packages("tidyterra")
install.packages("rnaturalearth")
install.packages("geodata")
```

### Load libraries:

```{r}
#| output: false
library(terra)
library(tidyverse)
library(rnaturalearth)
library(geodata)
library(sf)
library(tidyterra)
```

## Rasters

A raster is a spatial data structure that subdivides an extent into rectangles known as "cells" (or "pixels"). Each cell has the capacity to store one or more values. This type of data structure is commonly referred to as a "grid" and is frequently juxtaposed with simple features.

The terra package offers functions designed for the creation, reading, manipulation, and writing of raster data. The terra package is built around a number of "classes" of which the SpatRaster and SpatVector are the most important.

#### SpatRaster

A SpatRaster object stores a number of fundamental parameters that describe it. These include the number of columns and rows, the coordinates of its spatial extent, and the coordinate reference system. In addition, a SpatRaster can store information about the file(s) in which the raster cell values are stored.

#### SpatVector

A SpatVector represents "vector" data, that is, points, lines or polygon geometries and their tabular attributes.

#### Working with climate data:

To become familiar with working with rasters, we will download climate data for an area of interest (AOI).

```{r}
# First create an AOI
world <- ne_countries(scale = "medium", returnclass = "sf")
aoi_global <- world %>% filter(continent %in% c("Europe", "Asia", "North America", "South America", "Africa", "Oceania"))
```

Visualize your AOI:

```{r}
# Visualize it here
# ...
```

We will use WorldClim, a set of global climate layers at spatial resolutions from 30 arc-seconds (~1 km) down to 10 arc-minutes. Its values represent average monthly climate normals for 1970-2000 (Fick & Hijmans, 2017).

Download climate data using the geodata package. For this exercise we will use climate normals, multi-decadal averages for climate variables like temperature and precipitation. They provide a baseline that allows us to understand the location's average condition. geodata always downloads the full global raster, so we crop and mask it down to our AOI afterward.

You can access monthly precipitation ("prec"), monthly minimum temperature ("tmin"), monthly maximum temperature ("tmax") climate normals and more.

Download monthly precipitation (prec), then crop and mask it to the AOI:

```{r}
normals_ppt <- worldclim_global(var = "prec", res = 10, path = tempdir()) %>%
  crop(aoi_global) %>%
  mask(aoi_global)
```

What is this object:

```{r}
class(normals_ppt)
```

A SpatRaster with multiple layers holds one raster per time step (here, one per calendar month) that all share the same spatial extent and resolution.

To access the raster stack:

```{r}
normals_ppt
normals_ppt %>% names()

ggplot() + geom_spatraster(data = normals_ppt[[1]]) + scale_fill_gradient(na.value = "transparent")
```

What is the `+ scale_fill_gradient(na.value = "transparent")` doing?

#### Raster algebra

Many generic functions that allow for simple and elegant raster algebra have been implemented for SpatRaster objects, including the normal algebraic operators such as +, -, *, /, logical operators such as >, >=, <, ==, ! and functions such as abs, round, ceiling, floor, trunc, sqrt, log, log10, exp, cos, sin, max, min, range, prod, sum, any, all. In these functions you can mix terra objects with numbers, as long as the first argument is a terra object. If you use multiple SpatRaster objects, all objects must have the same resolution and origin.

Let's summarize monthly data to annual normals:

```{r}
normals_ppt_annual <- normals_ppt %>% sum(na.rm = TRUE)

# look at the object
normals_ppt_annual %>% plot()

# Check the name of the layers:
names(normals_ppt_annual)

# re-name the layers:
names(normals_ppt_annual) <- "ppt"
```

Summary functions (`min()`, `max()`, `mean()`, `prod()`, `sum()`, `median()`, `cv()`, `range()`, `any()`, `all()`) always return a SpatRaster object.

Use global if instead of a SpatRaster you want a single number summarizing the cell values of each layer.

```{r}
normals_ppt_annual %>% global(na.rm = TRUE, mean)
```

#### Spatial Summaries

You might also find it useful to create zonal summaries for each polygon within the simple feature. To do this we can use the function zonal, which takes a SpatRast and a SpatVect. Since aoi_global has one feature per country, this gives us a country-level summary:

```{r}
normals_ppt_annual_country <- zonal(x = normals_ppt_annual,
                                     z = vect(aoi_global), fun = "mean", as.polygons = TRUE, na.rm = TRUE)
```

What did the function return?

```{r}
class(normals_ppt_annual_country)
```

Convert the SpatVect back to a simple feature and plot it.

```{r}
normals_ppt_annual_country_sf <- st_as_sf(normals_ppt_annual_country)

ggplot(data = normals_ppt_annual_country_sf) + geom_sf(aes(fill = ppt))
```

#### Extracting information to a point file:

For this exercise you will use your fluxnet_ch4_sf vector file you saved last week and the monthly precipitation data downloaded (normals_ppt).

Import your point file fluxnet_ch4_sf:

```{r}
fluxnet_ch4_sf <- st_read(dsn = "data/products", layer = "fluxnet_ch4")
```

Ensure both files have the same coordinate reference system (CRS):

```{r}
fluxnet_ch4_sf
normals_ppt
```

You can transform one of the files if they don't match. Here I transform the vector:

```{r}
fluxnet_ch4_sf <- st_transform(fluxnet_ch4_sf, crs = crs(normals_ppt))
```

To check to see if everything lines up, I plot the files together:

```{r}
ggplot() + geom_spatraster(data = normals_ppt[[1]]) + geom_sf(data = fluxnet_ch4_sf) + scale_fill_gradient(na.value = "transparent")
```

Extract information from your raster stack using `terra::extract()`

```{r}
fluxnet_ch4_ppt <- terra::extract(normals_ppt, fluxnet_ch4_sf)
head(fluxnet_ch4_ppt)
```

What did the extract function return?

```{r}
class(fluxnet_ch4_ppt)
```

Combine extracted information to the simple feature:

```{r}
fluxnet_ch4_ppt_sf <- fluxnet_ch4_sf %>% cbind(fluxnet_ch4_ppt)
```

Visualize your work. Each extracted column is named after its WorldClim layer (e.g. `wc2.1_10m_prec_01`), so look it up with `names()` instead of typing it in by hand:

```{r}
names(fluxnet_ch4_ppt_sf)

ppt_col <- names(fluxnet_ch4_ppt_sf)[grepl("prec", names(fluxnet_ch4_ppt_sf))][1]

ggplot() + geom_sf(data = aoi_global) + geom_sf(data = fluxnet_ch4_ppt_sf, aes(col = .data[[ppt_col]]))
```

FLUXNET data can be used to understand patterns in natural methane fluxes. Evaluating the conditions where measurements are taken is essential to designing a useful model.

## Download monthly temperature (tmin and tmax) to understand the differences in temperature for the tower locations. Use precipitation and temperature summaries in the assignment.

## Assignment

We will use data from FLUXNET CH4 to explore patterns in natural methane emissions. Explore the distribution of tower sites and create visualizations that may be helpful to understand in the design and development of models. You are welcome to use any additional data or just new plot types.

#### References

Fick, S.E. and R.J. Hijmans, 2017. WorldClim 2: new 1km spatial resolution climate surfaces for global land areas. International Journal of Climatology 37 (12): 4302-4315. <https://doi.org/10.1002/joc.5086>
