---
title: "Raster Data: Climate Exposure and Tower Similarity"
editor:
  mode: source
---

## Mapping climate surfaces

Vector data told us where FLUXNET CH4 towers are located. Raster data lets us ask what environmental conditions surround those towers. In this workshop, we will use gridded climate layers to evaluate the climate exposure represented by methane tower sites and to identify towers that may be useful analogs for thinking about methane fluxes in Yale-Myers Forest wetlands.

We will ask:

> What climate conditions do FLUXNET CH4 towers represent, and which towers are most similar to Yale-Myers Forest?

By the end of this workshop, you will be able to:

1.  Read and inspect raster data with `terra`.
2.  Understand `SpatRaster` and `SpatVector` objects.
3.  Crop and mask rasters to an area of interest.
4.  Use raster algebra to create new climate layers.
5.  Summarize raster values globally and by polygon zones.
6.  Extract raster values to point locations.
7.  Compare tower-site climate exposure to gridded climate space.
8.  Calculate environmental dissimilarity among tower sites.
9.  Use multidimensional scaling (MDS) to visualize which towers are more or less similar.

### Install libraries for this workshop

```{r}
#| eval: false
install.packages("sf")
install.packages("terra")
install.packages("tidyverse")
install.packages("rnaturalearth")
```

### Load libraries

```{r}
#| output: false
library(terra)
library(tidyverse)
library(rnaturalearth)
library(sf)
```

# What is a raster?

A raster is a spatial data structure that subdivides an area into cells, often called pixels. Each cell stores one or more values. Climate rasters are useful because they let us describe environmental gradients everywhere, not only at point locations.

The `terra` package is built around two important classes:

-   `SpatRaster`: gridded raster data.
-   `SpatVector`: vector data such as points, lines, or polygons.

# Read climate rasters

This course includes a global climate raster with four layers:

-   `MAP`: mean annual precipitation in mm.
-   `TMIN`: mean minimum temperature in degrees C.
-   `TMAX`: mean maximum temperature in degrees C.
-   `MAT`: mean annual temperature in degrees C.

```{r}
global_climate <- rast("data/GlobalClimate.tif")
global_climate
```

Look at the layer names:

```{r}
names(global_climate)
```

Plot mean annual precipitation:

```{r}
plot(global_climate$MAP,
     main = "Mean annual precipitation")
```

Plot mean annual temperature:

```{r}
plot(global_climate$MAT,
     main = "Mean annual temperature")
```

# Raster algebra

Raster algebra lets us create new layers from existing layers. Temperature range is a simple measure of climate exposure:

```{r}
temp_range <- global_climate$TMAX - global_climate$TMIN
names(temp_range) <- "temp_range"
```

Combine the original climate layers with the new temperature range layer:

```{r}
climate_stack <- c(global_climate, temp_range)
names(climate_stack)
```

Plot the new raster:

```{r}
plot(climate_stack$temp_range,
     main = "Temperature range")
```

Use `global()` when you want one summary value for an entire raster layer:

```{r}
global(climate_stack, fun = "mean", na.rm = TRUE)
```

# Crop and mask to an area of interest

Cropping trims a raster to a rectangular extent. Masking removes cells outside an area of interest. Here we use countries from `rnaturalearth` to create a South America polygon and summarize climate over that region.

```{r}
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(continent %in% c("Europe", "Asia", "North America",
                          "South America", "Africa", "Oceania")) %>%
  st_make_valid()

south_america <- world %>%
  filter(continent == "South America") %>%
  st_union() %>%
  st_as_sf()
```

Crop and mask mean annual precipitation to South America:

```{r}
south_america_ppt <- climate_stack$MAP %>%
  crop(south_america) %>%
  mask(south_america)

plot(south_america_ppt,
     main = "Mean annual precipitation in South America")
```

What did the crop and mask operations return?

```{r}
class(south_america_ppt)
south_america_ppt
```

# Zonal summaries

Zonal summaries calculate raster statistics inside polygon zones. Here we calculate mean annual precipitation for each country in South America. Keeping the example regional makes the computation faster while preserving the same workflow.

```{r}
south_america_countries <- world %>%
  filter(continent == "South America")

country_ppt <- zonal(x = south_america_ppt,
                     z = vect(south_america_countries),
                     fun = "mean",
                     as.polygons = TRUE,
                     na.rm = TRUE)
```

What did `zonal()` return?

```{r}
class(country_ppt)
```

Convert the result back to an `sf` object and map it:

```{r}
country_ppt_sf <- st_as_sf(country_ppt)

ggplot(data = country_ppt_sf) +
  geom_sf(aes(fill = MAP), color = NA) +
  scale_fill_viridis_c(na.value = "transparent") +
  labs(title = "Country-level mean annual precipitation in South America",
       fill = "MAP (mm)")
```

Which South American countries have the highest mean annual precipitation in this dataset?

```{r}
country_ppt_sf %>%
  st_drop_geometry() %>%
  select(name, continent, MAP) %>%
  arrange(desc(MAP)) %>%
  head()
```

# Connecticut and Yale-Myers Forest

This course also includes Connecticut climate rasters for 2021. We can use them to practice with multi-layer rasters and to extract climate information for Yale-Myers Forest.

```{r}
ct_ppt_2021 <- rast("data/TERRA_PPT_2021_CT.tif")
ct_tmean_2021 <- rast("data/TERRA_TMEAN_2021_CT.tif")
```

Each raster has one layer per month. Keep the 2021 months:

```{r}
ct_ppt_2021 <- ct_ppt_2021[[1:12]]
ct_tmean_2021 <- ct_tmean_2021[[1:12]]

names(ct_ppt_2021)
names(ct_tmean_2021)
```

Summarize monthly Connecticut layers to annual precipitation and mean temperature:

```{r}
ct_annual_ppt_2021 <- sum(ct_ppt_2021, na.rm = TRUE)
names(ct_annual_ppt_2021) <- "ppt_2021"

ct_mean_tmean_2021 <- mean(ct_tmean_2021, na.rm = TRUE)
names(ct_mean_tmean_2021) <- "tmean_2021"
```

Plot the Connecticut rasters:

```{r}
plot(ct_annual_ppt_2021,
     main = "Connecticut 2021 annual precipitation")

plot(ct_mean_tmean_2021,
     main = "Connecticut 2021 mean temperature")
```

Create a Yale-Myers Forest point. These are the same coordinates used in the Data Basics workshop:

```{r}
yale_myers_sf <- tibble(
  SITE_ID = "Yale-Myers",
  SITE_NAME = "Yale-Myers Forest wetlands",
  IGBP = "WET",
  LOCATION_LAT = 41.952909,
  LOCATION_LONG = -72.123859
) %>%
  st_as_sf(coords = c("LOCATION_LONG", "LOCATION_LAT"),
           crs = 4326,
           remove = FALSE)
```

Extract 2021 Connecticut climate values for Yale-Myers:

```{r}
yale_myers_ct_climate <- terra::extract(c(ct_annual_ppt_2021, ct_mean_tmean_2021),
                                        vect(yale_myers_sf))

yale_myers_ct_climate
```

Map Yale-Myers on the Connecticut precipitation raster:

```{r}
ct_annual_ppt_df <- as.data.frame(ct_annual_ppt_2021,
                                  xy = TRUE,
                                  na.rm = TRUE)

ggplot(data = ct_annual_ppt_df) +
  geom_raster(aes(x = x, y = y, fill = ppt_2021)) +
  geom_sf(data = yale_myers_sf,
          color = "red",
          size = 3,
          inherit.aes = FALSE) +
  scale_fill_viridis_c(na.value = "transparent") +
  labs(title = "Yale-Myers Forest on Connecticut 2021 precipitation",
       fill = "Precipitation (mm)")
```

# Extract raster climate to tower sites

Read the FLUXNET site table and create point locations for all sites with coordinates:

```{r}
fluxnet_sites <- read.csv("data/FluxNet_Sites_2024.csv")

fluxnet_sites_located <- fluxnet_sites %>%
  filter(!is.na(LOCATION_LONG), !is.na(LOCATION_LAT)) %>%
  mutate(site_group = if_else(FLUXNET.CH4 != "",
                              "CH4 site",
                              "Other FLUXNET site"))

fluxnet_sites_sf <- st_as_sf(fluxnet_sites_located,
                              coords = c("LOCATION_LONG", "LOCATION_LAT"),
                              crs = 4326,
                              remove = FALSE)
```

Extract the global climate layers to each tower point:

```{r}
tower_climate_values <- terra::extract(climate_stack,
                                       vect(fluxnet_sites_sf)) %>%
  select(-ID) %>%
  rename_with(~ paste0("raster_", .x))

tower_climate <- fluxnet_sites_sf %>%
  st_drop_geometry() %>%
  bind_cols(tower_climate_values)
```

Look at the extracted values:

```{r}
tower_climate %>%
  select(SITE_ID, SITE_NAME, site_group, IGBP,
         raster_MAP, raster_MAT, raster_TMIN, raster_TMAX, raster_temp_range) %>%
  head()
```

Add Yale-Myers Forest to the extracted climate table:

```{r}
yale_myers_global_values <- terra::extract(climate_stack,
                                           vect(yale_myers_sf)) %>%
  select(-ID) %>%
  rename_with(~ paste0("raster_", .x))

yale_myers_climate <- yale_myers_sf %>%
  st_drop_geometry() %>%
  mutate(site_group = "Yale-Myers reference",
         FLUXNET.CH4 = "Yale-Myers reference") %>%
  bind_cols(yale_myers_global_values)

climate_context <- bind_rows(
  tower_climate,
  yale_myers_climate
)
```

# Raster-based representativeness

Compare CH4 tower climate conditions with all FLUXNET tower locations:

```{r}
ggplot(data = tower_climate,
       aes(x = raster_MAT, y = raster_MAP, color = site_group)) +
  geom_point(alpha = 0.75) +
  labs(title = "Climate space represented by FLUXNET tower sites",
       x = "Mean annual temperature (degrees C)",
       y = "Mean annual precipitation (mm)",
       color = "Site group")
```

Add Yale-Myers Forest:

```{r}
ggplot(data = climate_context,
       aes(x = raster_MAT, y = raster_MAP, color = site_group)) +
  geom_point(alpha = 0.75) +
  geom_point(data = yale_myers_climate,
             color = "red",
             size = 4) +
  labs(title = "Yale-Myers Forest in raster-derived climate space",
       x = "Mean annual temperature (degrees C)",
       y = "Mean annual precipitation (mm)",
       color = "Site group")
```

Compare tower locations with a sample of global raster cells:

```{r}
set.seed(730)

global_climate_sample <- spatSample(climate_stack[[c("MAT", "MAP")]],
                                    size = 1000,
                                    method = "regular",
                                    na.rm = TRUE,
                                    as.df = TRUE) %>%
  select(MAT, MAP)

ggplot() +
  geom_point(data = global_climate_sample,
             aes(x = MAT, y = MAP),
             color = "gray85",
             alpha = 0.3) +
  geom_point(data = tower_climate %>% filter(site_group == "CH4 site"),
             aes(x = raster_MAT, y = raster_MAP, color = IGBP),
             alpha = 0.85) +
  geom_point(data = yale_myers_climate,
             aes(x = raster_MAT, y = raster_MAP),
             color = "red",
             size = 4) +
  labs(title = "CH4 towers and Yale-Myers relative to sampled global climate space",
       subtitle = "Gray points are sampled raster cells",
       x = "Mean annual temperature (degrees C)",
       y = "Mean annual precipitation (mm)",
       color = "IGBP")
```

# Environmental dissimilarity among CH4 towers

To identify which towers are more or less similar to Yale-Myers Forest, we can compare sites using their raster-derived climate variables. We will use standardized Euclidean distance so variables with large units, such as precipitation, do not dominate the comparison only because of scale.

For this first-pass analysis, we will compare sites using:

-   `raster_MAP`;
-   `raster_MAT`;
-   `raster_TMIN`;
-   `raster_TMAX`;
-   `raster_temp_range`.

```{r}
climate_vars <- c("raster_MAP", "raster_MAT", "raster_TMIN",
                  "raster_TMAX", "raster_temp_range")

analog_candidates <- climate_context %>%
  filter(site_group %in% c("CH4 site", "Yale-Myers reference")) %>%
  select(SITE_ID, SITE_NAME, IGBP, site_group, all_of(climate_vars)) %>%
  drop_na()
```

Scale the variables and calculate distances:

```{r}
scaled_climate <- analog_candidates %>%
  select(all_of(climate_vars)) %>%
  scale()

climate_distance <- dist(scaled_climate)
```

Calculate each CH4 tower's distance from Yale-Myers Forest:

```{r}
yale_index <- which(analog_candidates$SITE_ID == "Yale-Myers")

distance_to_yale <- as.matrix(climate_distance)[yale_index, ]

yale_myers_analogs <- analog_candidates %>%
  mutate(distance_to_yale = as.numeric(distance_to_yale)) %>%
  filter(SITE_ID != "Yale-Myers") %>%
  arrange(distance_to_yale)
```

Which CH4 towers are most similar to Yale-Myers based on these raster climate variables?

```{r}
yale_myers_analogs %>%
  select(SITE_ID, SITE_NAME, IGBP,
         raster_MAP, raster_MAT, raster_temp_range, distance_to_yale) %>%
  head(10)
```

Because Yale-Myers wetlands are wetlands, it may be useful to separately inspect wetland towers:

```{r}
yale_myers_analogs %>%
  filter(IGBP == "WET") %>%
  select(SITE_ID, SITE_NAME, IGBP,
         raster_MAP, raster_MAT, raster_temp_range, distance_to_yale) %>%
  head(10)
```

::: {.callout-note}
## Interpret analogs carefully
This is a climate-similarity analysis, not proof that a tower is a perfect analog for Yale-Myers wetlands. Methane fluxes also depend on water table depth, wetland type, soils, vegetation, disturbance history, and measurement design. The goal is to identify towers that may be useful starting points for comparison.
:::

# MDS visualization

Multidimensional scaling (MDS) takes a distance matrix and places sites into a low-dimensional space. Sites that plot closer together are more similar based on the variables used to calculate the distance.

```{r}
mds <- cmdscale(climate_distance, k = 2)

mds_scores <- analog_candidates %>%
  bind_cols(as.data.frame(mds) %>%
              rename(MDS1 = V1, MDS2 = V2))
```

Label Yale-Myers and the five most similar CH4 towers:

```{r}
label_sites <- mds_scores %>%
  filter(SITE_ID == "Yale-Myers" |
           SITE_ID %in% yale_myers_analogs$SITE_ID[1:5])
```

Plot the MDS:

```{r}
ggplot(data = mds_scores,
       aes(x = MDS1, y = MDS2)) +
  geom_point(aes(color = site_group, shape = IGBP == "WET"),
             alpha = 0.75,
             size = 2.5) +
  geom_text(data = label_sites,
            aes(label = SITE_ID),
            nudge_y = 0.12,
            size = 3) +
  labs(title = "MDS of CH4 tower climate similarity",
       subtitle = "Distance calculated from raster-derived climate variables",
       x = "MDS axis 1",
       y = "MDS axis 2",
       color = "Site group",
       shape = "Wetland tower")
```

::: {.callout-tip}
## Course product
This workshop produces a raster evidence brief and analog tower table. Use the standard deliverable shape from the Final Project Roadmap: one clear question, one reproducibility note, at least one table, one or two figures, one interpretation paragraph, and one limitation or next-step paragraph.
:::

# Assignment

Create a short raster evidence brief that evaluates climate exposure and tower similarity for Yale-Myers Forest wetlands.

Your report should answer:

1.  What precipitation and temperature conditions are represented by FLUXNET CH4 tower sites?
2.  Where does Yale-Myers Forest fall in raster-derived climate space?
3.  Which CH4 towers are most climatically similar to Yale-Myers Forest?
4.  Which wetland CH4 towers are most similar to Yale-Myers Forest?
5.  Based on the MDS plot and distance table, which towers would you prioritize as analogs for estimating or contextualizing methane fluxes from Yale-Myers wetlands?
6.  What important wetland controls are missing from this raster-only similarity analysis?

Include at least:

-   one raster map;
-   one climate-space plot;
-   one distance or MDS figure;
-   one table of candidate analog towers.

#### References

Fick, S.E. and R.J. Hijmans, 2017. WorldClim 2: new 1km spatial resolution climate surfaces for global land areas. International Journal of Climatology 37 (12): 4302-4315. <https://doi.org/10.1002/joc.5086>
